---
title: "Data Assembly: 2009 NHTS (School Trips)"
author: "Aanchal Chopra"
date: "`r Sys.Date()`"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Data Assembly notes

## Constructing the Sample

Data for this analysis is drawn from the 2009 National Household Travel Survey (https://nhts.ornl.gov/).
  
#### Criteria for inclusion in sample

**Data**

* Data not missing for any outcome or predictor variables

**Trip (to ensure one trip per child, and only school trips):**

* Trip distance is shorter than 2km/1.25 miles
* Trip ends at school
* Trip is not by transit, motorcycle, or an unspecified mode (these are rare and I assume they are not available to the remaining children)
* Child does not use a school bus for the trip to _or_ from school (assume the remaining students are ineligible for school bus service)
* Survey does not indicate that the child drove to school unaccompanied (assume these are survey coding errors)
* Survey does not indicate the child traveled to school by car with "only siblings" unless the child has a sibling who is a driver.
* Trip does not begin AND end at school
* Trip ends before 10am
* This is the first qualifying trip of the day

**Traveler**

* Traveler is between 8 and 13

#### Assembling Multi-stage trips

If the trip ending at school begins with a transfer from another mode, the prior trip is included as part of this trip, and trip characteristics are determined as follows:
  
* Matched with the characteristics of the longest-distance segment of the trip:
  * Mode
  * Presence of others
*Summed across all segments:
  * Trip distance
*Taken from the first segment of the trip:
  * Population density at trip origin
  * All individual- and household-level variables
* Taken from last segment of trip
  * Population density at trip destination
  
## Defining the variables

#### Outcome variables

* **'mode'**: One of:
  * 7 = car
  * 8 = walk
  * 9 = bike
  
* **'independence'**: (string) For purposes of this analysis, we describe all female household adults as moms, all male household adults as dads, and all household children as siblings. Also note that NHTS codes all household members as either male or female. We know how many non-household members are on a trip, but we don't know their ages, genders, or drivers status.The full independence variable takes one of the following 6 values to describe who was with the child on their trip to school:
  * 10 = alone: There was only one person (the child) on the trip
  * 21 = with mom and dad: The child was accompanied by a male household adult _and_ a female household adult
  * 22 = with mom: The child was accompanied by a female household adult but no male household adult
  * 23 = with dad: The child was accompanied by a male household adult but no female household adult
  * 24 = with non-household: The child was accompanied by non-household members and no household members were on the trip
  * 30 = with sibling: The child was accompanied by household children, but no household adults or non-household members
  
* **'ind_3'**: A simplified independence variable. One of:
  * 10 = alone: Same as 1 (alone) in the full independence variable
  * 20 = with adults: Combination of these values from the full independence variable
     * 21 (with mom and dad)
     * 22 (with mom)
     * 23 (with dad)
     * 24 (with non-household)
  * 30 = with kids: same as 30 (with sibling) in the full independence variable
  
* **'ind_3_alt'**: Same as 'ind_3a', but trips with non-household members (independence = 30) are classified as trips with kids. Since we don't know if the non-household members are kids or adults, we might want to test it both ways and see if it effects the result

* **'mode_ind'**: Combination of mode and the full independence variable. Takes the following values:
  * 721 = car with mom and dad
  * 722 = car with mom
  * 723 = car with dad
  * 724 = car with non-household
  * 730 = car with sibling
  * 810 = walk alone
  * 821 = walk with mom and dad
  * 822 = walk with mom
  * 823 = walk with dad
  * 824 = walk with non-household
  * 830 = walk with sibling
  * 910 = bike alone
  * 921 = bike with mom and dad
  * 922 = bike with mom
  * 923 = bike with dad
  * 924 = bike with non-household
  * 930 = bike with sibling
  
* **'mode_ind_3'**: Combination of mode and the simplified independence
  variable. Takes the following values
  * 720 = car with adult
  * 730 = car with kid
  * 810 = walk alone
  * 820 = walk with adults
  * 830 = walk with kids
  * 910 = bike alone
  * 920 = bike with adults
  * 930 = bike with kids
  
* **'mode_ind_3_alt'**: Combination of mode and the simplified independence variable that classifies non-household members as kids. Same categories as 'mode_ind_3.
  
#### Availability variables

**'av_car', 'av_walk', and 'av_bike'** indicate the trips for which travel by car, walking, or bike is available. We are assuming that these three modes are available for all children in the sample (even if there is not car in the household, since some children in the sample in zero-vehicle households _do_ travel by car) so this value is set to one for all cases.

The following variables indicate the availability of independence alternatives:
  
  * alone_avail: TRUE/1 for all trips
  * with_mom_dad_avail: True if there is both a female and a male adult in the household.
  * with_mom_avail: True if there is a female adult in the household
  * with_dad_avail: True if there is a male adult in the household
  * with_non_hh_avail: True for all trips
  * with_sib_avail: True if there are any other children in the household
  * with_adult_avail: True for all trips
  
#### Predictor variables
  
**Household-level variables**:
  
  * income_k: NHTS codes income in one of 11 income categories. We convert this to a continuous variable by assigning households in each category the mid-point value of that category. The highest income category is for incomes greater than $200,000 per year. We assign an income of $250,000 to that category. log_income_k is the natural log of income_k.
  * veh_per_driver: We divide the number of household vehicles by the number of household drivers. We assign a value of zero to households with zero drivers
  * n_adults: The number of household adults
  * has_mom: A binary variable indicating whether there is a female adult in the household
  * has_dad: A binary variable indicating whether there is a male adult in the household
  * non_work_mom: A binary variable indicating whether there is a female adult in the household who is not a worker
  * non_work_dad: A binary variable indicating whether there is a male adult in the household who is not a worker
  
**Individual-level variables**:
  
  * age: The child's age
  *female: A binary variable indicating whether the child is female
  * has_lil_sib: A binary variable indicating whether there are any younger children in the household (includes children who are the same age as the respondent)
  * has_big_sib: A binary variable indicating whether there are any older children in the household
  
**Trip-level variables**:

  * distance: Trip distance in kilometers. The NHTS records distance in miles and these are converted to kilometers by multiplying by 1.609
    * log_distance is the natural log of distance
  * density: The approximate population density of the census block in which the trip begins or ends (whichever is higher). NHTS reports this value in people per square mile. We convert to people per square kilometer by dividing by 2.59. 
    * log_density is the natural log of density
    
# Data Assembly

#### 1 Load NHTS trip and person files
  
```{r, message = FALSE, warning = FALSE}

library(tidyverse)
library(knitr)
library(tidyr)
library(dplyr)
library(here)
library(kableExtra)
library(magrittr)
library(downloader)
library(naniar)

trips09 <- here("nhts",
              "data2009", 
              "DAYV2PUB.csv") |>
  read_csv(show_col_types = FALSE)

people09 <- here("nhts",
               "data2009", 
               "PERV2PUB.csv") |>
  read_csv(show_col_types = FALSE) |>
  mutate(person_hh = paste(PERSONID, HOUSEID, sep = "-"))

```

#### 2 Assemble ages of household members

```{r, message = FALSE, warning = FALSE}

hh_ages09 <- people09 |>
  select(HOUSEID, PERSONID, R_AGE) |>
  pivot_wider(names_from = PERSONID,
              names_prefix = "age_",
              values_from = R_AGE)
```

#### 3 Assemble relationships of all individuals

```{r, message = FALSE, warning = FALSE}

relationships09 <- people09 |>
  mutate(adult = R_AGE > 17,
         kid_age = ifelse(R_AGE < 18, R_AGE, -2),
         R_AGE = ifelse(R_AGE < 0, 999, R_AGE),
         non_worker_mom = R_AGE > 17 & WORKER != "01" & R_SEX == "02",
         non_worker_dad = R_AGE > 17 & WORKER != "01" & R_SEX != "02") |>
  group_by(HOUSEID) |>
  mutate(n_adults = sum(adult),
         num_records = n(),
         non_work_mom = sum(non_worker_mom) > 0,
         non_work_dad = sum(non_worker_dad) > 0) |>
  mutate(youngest_kid = ifelse(num_records > HHSIZE, 1, min(R_AGE)),
         oldest_kid = max(kid_age),
         n_children = HHSIZE - n_adults) |>
  mutate(has_big_sib = n_children > 1 & R_AGE != oldest_kid) |>
  mutate(has_lil_sib = n_children > 1 & (R_AGE != youngest_kid | !has_big_sib)) |>
  mutate(person_hh = paste(PERSONID, HOUSEID, sep = "-")) |>
  ungroup() |>
  select(person_hh, 
         has_lil_sib, 
         has_big_sib,
         non_work_mom,
         non_work_dad)
```

#### 4 Assemble trips and school trips

```{r, message = FALSE, warning = FALSE}

school_trips09 <- trips09 |>
  mutate(include_trip = (WHYTO == "21" &
                           WHYFROM != "21" & ## Updated coding to match 2009 variable
                           R_AGE > 7 & 
                           R_AGE < 14 &
                           as.numeric(ENDTIME) < 1000 &
                           TRPMILES > 0)) |>  ## Removed transfer variable
  mutate(mode = case_when(TRPTRANS == "11" ~ 1, # school bus ## Updated coding to match 2009 variables
                          TRPTRANS == "01" ~ 7, # car
                          TRPTRANS == "02" ~ 7,
                          TRPTRANS == "03" ~ 7,
                          TRPTRANS == "04" ~ 7,
                          TRPTRANS == "23" ~ 8, # walk
                          TRPTRANS == "22" ~ 9, # bike
                          TRPTRANS == "9" ~ 2,  # transit
                          TRPTRANS == "10" ~ 2,
                          TRPTRANS == "12" ~ 2, 
                          TRPTRANS == "13" ~ 2,
                          TRPTRANS == "14" ~ 2,
                          TRPTRANS == "15" ~ 2,
                          TRPTRANS == "16" ~ 2,
                          TRPTRANS == "17" ~ 2,
                          TRPTRANS == "18" ~ 2,
                          TRPTRANS == "24" ~ 2,
                          TRPTRANS == "07" ~ 3, # motorcycle
                          TRPTRANS == "97" ~ 4, # unspecified
                          TRUE ~ 5))|> ## Removed Transfers
  mutate(trip_person_hh = paste(TDTRPNUM, PERSONID, HOUSEID, sep = "-")) |>
  mutate(person_hh = paste(PERSONID, HOUSEID, sep = "-")) |>
  group_by(person_hh) |>
  mutate(can_sch_bus = sum(mode == 1) > 0) |>
  ungroup() |>
  filter(TRPMILES < 1.25) |>
  filter(include_trip > 0) |>
  filter(!duplicated(person_hh)) |>
  filter(mode > 6,
         !can_sch_bus) |>
  left_join(hh_ages) |>
  left_join(hh_genders) |>
  mutate(alone = NUMONTRP == 1,
         with_parent = 
           (ONTD_P1 == 1 & age_01 > 17) |
           (ONTD_P2 == 1 & age_02 > 17) |
           (ONTD_P3 == 1 & age_03 > 17) |
           (ONTD_P4 == 1 & age_04 > 17) |
           (ONTD_P5 == 1 & age_05 > 17) |
           (ONTD_P6 == 1 & age_06 > 17) |
           (ONTD_P7 == 1 & age_07 > 17) |
           (ONTD_P8 == 1 & age_08 > 17) |
           (ONTD_P9 == 1 & age_09 > 17) |
           (ONTD_P10 == 1 & age_10 > 17) |
           (ONTD_P11 == 1 & age_11 > 17) |
           (ONTD_P12 == 1 & age_12 > 17) |
           (ONTD_P13 == 1 & age_13 > 17),
         hh_only = NUMONTRP == NONHHCNT+1) |>
  left_join(relationships) |>
  mutate(independence = case_when(with_mom & with_dad ~ 21,
                                  with_mom ~ 22,
                                  with_dad ~ 23,
                                  alone ~ 10,
                                  NONHHCNT == 0 ~ 30,
                                  TRUE ~ 24)) |>
  mutate(ind_3 = ifelse(independence > 20 & independence < 30, 
                        20, independence)) |>
  mutate(ind_3_alt = case_when(independence > 10 & independence < 24 ~ 20,
                               independence == 24 ~ 30,
                               TRUE ~ independence)) |>
  filter(!(mode == 7 & independence == 10),
         !(mode == 7 & independence == 30 & !driver_sib)) |>
  mutate(mode_ind = mode * 100 + independence,
         mode_ind_3 = mode * 100 + ind_3,
         mode_ind_3_alt = mode * 100 + ind_3_alt) |>
  mutate(veh_per_driver = ifelse(DRVRCNT > 0, HHVEHCNT/DRVRCNT, 0)) |>
  mutate(income_k = case_when(HHFAMINC == "01" ~ 5,
                              HHFAMINC == "02" ~ 12.5,
                              HHFAMINC == "03" ~ 20,
                              HHFAMINC == "04" ~ 30,
                              HHFAMINC == "05" ~ 42.5,
                              HHFAMINC == "06" ~ 62.5,
                              HHFAMINC == "07" ~ 87.5,
                              HHFAMINC == "08" ~ 112.5,
                              HHFAMINC == "09" ~ 137.5,
                              HHFAMINC == "10" ~ 175,
                              HHFAMINC == "11" ~ 250,
                              TRUE ~ -9)) |>
  mutate(income_k = ifelse(income_k == 250, income_k, income_k * 1.14)) |> ## adjust for 2017 income, except if 250
  filter(income_k > 0) |>
  mutate(max_od_dens = HBPPOPDN) |> ## adjust density variable
  filter(max_od_dens > 0) |>
  rename(age = R_AGE) |>
  mutate(female = as.numeric(R_SEX == "02"),
         has_mom = as.numeric(has_mom),
         has_dad = as.numeric(has_dad),
         non_work_mom = as.numeric(non_work_mom),
         non_work_dad = as.numeric(non_work_dad),
         has_lil_sib = as.numeric(has_lil_sib),
         has_big_sib = as.numeric(has_big_sib),
         distance = TRPMILES * 1.609,
         density = max_od_dens / 2.59) |>
  mutate(with_mom_avail = ifelse(has_mom == 1, 1, 0),
         with_dad_avail = ifelse(has_dad ==1, 1, 0),
         with_mom_dad_avail = ifelse(has_mom + has_dad == 2, 1, 0),
         with_sib_avail = ifelse(has_lil_sib + has_big_sib > 0, 1, 0)) |>
  mutate(log_income_k = log(income_k),
         log_distance = log(distance),
         log_density = log(density),
         av_car = 1,
         av_walk = 1,
         av_bike = 1,
         alone_avail = 1,
         with_non_hh_avail = 1,
         with_adult_avail = 1) |>
  select(mode,
         independence,
         ind_3,
         ind_3_alt,
         mode_ind,
         mode_ind_3,
         mode_ind_3_alt,
         income_k,
         log_income_k,
         veh_per_driver,
         n_adults,
         non_work_mom,
         non_work_dad,
         age,
         female,
         has_lil_sib,
         has_big_sib,
         distance,
         log_distance,
         density,
         log_density,
         av_car,
         av_walk,
         av_bike,
         alone_avail,
         with_mom_dad_avail,
         with_mom_avail,
         with_dad_avail,
         with_non_hh_avail,
         with_sib_avail,
         with_adult_avail)
```

#### 5 Add column for year

```{r, message = FALSE, warning = FALSE}

school_trips <- school_trips |>
  mutate(year = 2009)

```

# Summary Statistics

#### Sample Size

```{r, message = FALSE, warning = FALSE, results='asis', echo=FALSE}

tibble(Unit = c("Households",
                "Children",
                "Trips"),
       Sample = c(nrow(school_trips),
                  nrow(school_trips),
                  nrow(school_trips))) |>
  kable(caption = "Sample size")|>
  kable_styling(full_width = FALSE, position = "left")

```

#### Outcome 1: Full Independence Variable

```{r,message = FALSE, warning = FALSE, results='asis', echo=FALSE}

 mode_ind_table_count <- school_trips |>
  group_by(independence, mode) |>
  summarise(n = n()) |>
  ungroup() |>
  pivot_wider(names_from = mode, values_from = n) |>
  replace_na(list(`7` = 0)) |>
  cbind(`-` = c("Alone", 
                  "With mom and dad", 
                  "With mom",
                  "With dad",
                  "With non-household",
                  "With siblings")) |>
  rename(Car = `7`,
         Walk = `8`,
         Bike = `9`) |>
  select(`-`,
         Car,
         Bike, 
         Walk) |>
  mutate(Total = Car + Bike + Walk) |>
  rbind(tibble(`-` = "Total",
                  Car = sum(school_trips$mode == 7),
                  Walk = sum(school_trips$mode == 8),
                  Bike = sum(school_trips$mode ==9),
                  Total = nrow(school_trips)))

mode_ind_table_pct <- mode_ind_table_count |>
  mutate(across(-`-`, ~ paste0(round(.x * 100/ nrow(school_trips),1), "%")))

kable(mode_ind_table_count,
      caption = "Number of trips in sample by mode and (full) independence") |>
kable_styling(full_width = FALSE, position = "left")

kable(mode_ind_table_pct,
      caption = "Share of trips in sample by mode and (full) independence")  |>
kable_styling(full_width = FALSE, position = "left")

```  

#### Outcome 2: Simplified Independence Variable

```{r,message = FALSE, warning = FALSE, results='asis', echo=FALSE}

mode_ind3_table_count <- school_trips |>
  group_by(ind_3, mode) |>
  summarise(n = n()) |>
  ungroup() |>
  pivot_wider(names_from = mode, values_from = n) |>
  replace_na(list(`7` = 0)) |>
  cbind(`-` = c("Alone", 
                "With adults",
                "With kids")) |>
  rename(Car = `7`,
         Walk = `8`,
         Bike = `9`) |>
  select(`-`,
         Car,
         Bike, 
         Walk) |>
  mutate(Total = Car + Bike + Walk) |>
  rbind(tibble(`-` = "Total",
               Car = sum(school_trips$mode == 7),
               Walk = sum(school_trips$mode == 8),
               Bike = sum(school_trips$mode ==9),
               Total = nrow(school_trips))) 

mode_ind3_table_pct <- mode_ind3_table_count |>
  mutate(across(-`-`, ~ paste0(round(.x * 100/ nrow(school_trips),1), "%")))

kable(mode_ind3_table_count,
      caption = "Number of trips in sample by mode and (simplified) independence")|>
kable_styling(full_width = FALSE, position = "left")

kable(mode_ind3_table_pct,
      caption = "Share of trips in sample by mode and (simplified) independence")|>
kable_styling(full_width = FALSE, position = "left")

```

#### Outcome 3: Alternative Simplified Independence Variable

```{r,message = FALSE, warning = FALSE, results='asis', echo=FALSE}

mode_ind3alt_table_count <- school_trips |>
  group_by(ind_3_alt, mode) |>
  summarise(n = n()) |>
  ungroup() |>
  pivot_wider(names_from = mode, values_from = n) |>
  replace_na(list(`7` = 0)) |>
  cbind(`-` = c("Alone", 
                "With adults",
                "With kids")) |>
  rename(Car = `7`,
         Walk = `8`,
         Bike = `9`) |>
  select(`-`,
         Car,
         Bike, 
         Walk) |>
  mutate(Total = Car + Bike + Walk) |>
  rbind(tibble(`-` = "Total",
               Car = sum(school_trips$mode == 7),
               Walk = sum(school_trips$mode == 8),
               Bike = sum(school_trips$mode ==9),
               Total = nrow(school_trips))) 

mode_ind3alt_table_pct <- mode_ind3alt_table_count |>
  mutate(across(-`-`, ~ paste0(round(.x * 100/ nrow(school_trips),1), "%")))

kable(mode_ind3alt_table_count,
      caption = "Number of trips in sample by mode and (alternative simplified) independence")|>
kable_styling(full_width = FALSE, position = "left")

kable(mode_ind3alt_table_pct,
      caption = "Share of trips in sample by mode and (alternative simplified) independence")|>
kable_styling(full_width = FALSE, position = "left")

```

#### Choice Availability

```{r,message = FALSE, warning = FALSE, results='asis', echo=FALSE}

tibble(`Full independence variable` = c("Alone",
                                     "With mom and dad",
                                     "With mom",
                                     "With dad",
                                     "With non-household",
                                     "With siblings"),
    `Percent selected` = c(mean(school_trips$independence == 10),
                           mean(school_trips$independence == 21),
                           mean(school_trips$independence == 22),
                           mean(school_trips$independence == 23),
                           mean(school_trips$independence == 24),
                           mean(school_trips$independence == 30)),
    `Percent available` = c(1,
                            mean(school_trips$with_mom_dad_avail),
                            mean(school_trips$with_mom_avail),
                            mean(school_trips$with_dad_avail),
                            1,
                            mean(school_trips$with_sib_avail))) |>
  mutate(`Percent selected` = paste0(round(`Percent selected`*100), "%"),
         `Percent available` = paste0(round(`Percent available`*100), "%")) |>
  kable(caption = "Prevalence and availability of full independence choices")|>
  kable_styling(full_width = FALSE, position = "left")

```

#### Predictors

```{r,message = FALSE, warning = FALSE, results='asis', echo=FALSE}

school_trips |>
  select(-mode,
         -independence,
         -ind_3,
         -ind_3_alt,
         -mode_ind,
         -mode_ind_3,
         -mode_ind_3_alt,
         -av_bike,
         -av_car,
         -av_walk,
         -with_adult_avail,
         -with_dad_avail,
         -with_mom_avail,
         -with_mom_dad_avail,
         -with_non_hh_avail,
         -with_sib_avail) |>
  pivot_longer(cols = everything(),
               names_to = "Predictor",
               values_to = "Value") |>
  group_by(Predictor) |>
  summarise(Mean = mean(Value),
            `Standard Deviation` = ifelse(length(unique(Value)) == 2, 
                                          999,
                                          sd(Value))) |>
  replace_with_na(list(`Standard Deviation` = 999)) |>
  kable(digits = 3,
        caption = "Descriptive statistics of predictor variables")|>
  kable_styling(full_width = FALSE, position = "left")
```

# Export Data

```{r,message = FALSE, warning = FALSE, results='asis', echo=FALSE}
school_trips |>
  write_rds(file = here("data",
                        "only-school",
                        "usa-2009-schooltrips.rds"))
```
