---
title: "Initial Apollo Models"
author: "Macfarlane"
format: html
editor: visual
---

## Models

These are simple test models where the independence alternatives are `alone`, `parent` and `others` and the modes are `walk`, `bike`, and `drive`. Additionally a class membership model only uses the `female` dummy and the alternative utilities only have `n_adults`.

## Latent class MNL

```{r setup, include = FALSE}
library(apollo)
library(tidyverse)
library(here)
```

This model simply has a class membership model leading into one of two potential MNL models. There are actually a number of differences between the two sets of estimated coefficients. Need to work on a

```{r lcmnl}
lcmnlind3 <- read_rds(here("models/latent_class/LCMNLind3_model.rds"))
apollo_modelOutput(lcmnlind3)
```

## Nested Logit

This

```{r nl}
nlind3 <- read_rds(here("models/latent_class/NLind3_model.rds"))
apollo_modelOutput(nlind3)
```

## Latent class NL

```{r lcnl}
lcnlind3 <- read_rds(here("models/latent_class/LCNLind3_model.rds"))
apollo_modelOutput(lcnlind3)
```
