---
title: "All Trips_2009-2017-2022"
author: "Aanchal Chopra"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Required libraries

We're only using the `tidyverse`,  `here`, and `ggplot2` libraries.

```{r, message=FALSE}

library(tidyverse)
library(here)
library(ggplot2)
```

# Assemble person-level data

We need information on the ages of the household members on each trip to 
determine whether a child was accompanied by a household adult.

We also need to determine whether each household includes a non-working female
adults and/or a non-working male adult and the ages of the youngest and 
oldest children in each household (to determine if the trip-maker has a 
younger or older sibling).

## Load person-level data

We start by loading the person-level data files for three years, selecting the
same set of variables from each year, adding a variable to indicate which
year's sample the household is from, and renaming the sex variable to have the
same name in all three years.

Then we combine the three datasets for 2009, 2017 and 2022. 

```{r}

#import and filter 2017 person file
people17 <- here("nhts",
                 "data2017", 
                 "perpub.csv") |>
  read_csv(show_col_types = FALSE) |>
  select(HOUSEID, 
         PERSONID,
         R_AGE,
         R_SEX_IMP,
         WORKER,
         HHSIZE) |>
  mutate(year = 2017) |>
  rename(sex = R_SEX_IMP)

#import and filter 2009 person file
people09 <- here("nhts",
               "data2009", 
               "PERV2PUB.csv") |>
  read_csv(show_col_types = FALSE) |>
  select(HOUSEID, 
         PERSONID,
         R_AGE,
         R_SEX,
         WORKER,
         HHSIZE) |>
  mutate(year = 2009) |>
  rename(sex = R_SEX)

#import and filter 2022 person file
people22 <- here("nhts",
                 "data2022", 
                 "perv2pub.csv") |>
  read_csv(show_col_types = FALSE) |>
  select(HOUSEID, 
         PERSONID,
         R_AGE,
         R_SEX_IMP,
         WORKER,
         HHSIZE) |>
  mutate(year = 2022) |>
  rename(sex = R_SEX_IMP)
  
#combine three person files
people <- rbind(people09, people17, people22) |>
  mutate(person_hh_year = paste(PERSONID, HOUSEID, year, sep = "-")) |>
  mutate(hh_year = paste(HOUSEID, year, sep = "-"))

```

## Get ages of household members

We'll reshape the person-level data into a household-level dataset with
a column indicating the age of each person in the household.

```{r}
hh_ages <- people |>
  select(hh_year, PERSONID, R_AGE) |>
  pivot_wider(names_from = PERSONID,
              names_prefix = "age_",
              values_from = R_AGE)
```

## Get presence of non-working adult men and women and age ranges of children

For each household, we'll first determine which households include an adult 
(if age is greater than 17 i.e., 18 or above) or kid (age < 18 i.e., 17 or 
below). For adults, we assume all female adults are mothers and all male adults 
are fathers (or perform equivalent supervisory functions). 

Then, we determine which parents are working. We'll then reshape this 
household-level data to determine the presence of non-working adult men and 
women and the age of the oldest and youngest children.

```{r}
who_in_hh <- people |>
  mutate(adult = R_AGE > 17,
         kid_age = ifelse(R_AGE < 18, R_AGE, -2),
         R_AGE = ifelse(R_AGE < 0, 999, R_AGE),
         non_worker_mom = R_AGE > 17 & WORKER != "01" & sex == "02", #non-working parent is assumed to be an adult (>17) and not a Worker (01)
         non_worker_dad = R_AGE > 17 & WORKER != "01" & sex != "02") |>
  group_by(hh_year, HHSIZE) |>
  summarize(num_records = n(),
         n_adults = sum(adult),
         non_work_mom = sum(non_worker_mom) > 0,
         non_work_dad = sum(non_worker_dad) > 0,
         min_age = min(R_AGE),
         max_kid_age = max(kid_age)) |>
  mutate(youngest_kid = ifelse(num_records < HHSIZE, 1, min_age),
         oldest_kid = ifelse(max_kid_age > 0, max_kid_age, 1),
         n_children = HHSIZE - n_adults) |>
  ungroup() |>
  filter(n_children > 0) |>
  select(hh_year, 
         non_work_dad,
         non_work_mom,
         oldest_kid, 
         youngest_kid)
```
## Trips data

We then load in the trip-level data files for the three years, adding a variable
to indicate the year of the sample.

In 2017 and 2022, some trips include a 'transfer' variable where the individual
has changed mode of transport. In these cases, we combine the legs of the journey
to create a single trip where:
- Mode = the mode used for the longest leg of the journey
- Trip distance = sum of all legs of the journey
- Endtime = endtime of the last leg of the journey

We assume there are no transfer trips in 2009 as this variable does not exist.

### Import 2009 trips data

```{r}

#import 2009 trips data and add a column indicating year
trips09 <- here("nhts",
              "data2009", 
              "DAYV2PUB.csv") |>
  read_csv(show_col_types = FALSE) |>
  mutate(year = 2009)

```

### Combine 2017 transfer trips 

```{r}

#import 2017 trips data and add a column indicating year
trips17 <- here("nhts",
                "data2017", 
                "trippub.csv") |>
  read_csv(show_col_types = FALSE) |>
  mutate(year = 2017)|>
  arrange(HOUSEID, PERSONID, ENDTIME)

#generate a version of 2017 trips without transfer trips  
trips17_excl_transfers <- trips17 |>
  mutate(HOUSE_PERSONID = paste(HOUSEID, PERSONID, sep = "-")) |>
  mutate(transfer = ifelse(WHYFROM == "07" | WHYTO == "07", 1, 0)) |>
  filter(transfer == 0) |>
  select(-c("transfer"))

#generate a version of 2017 trips with only the trips involving transfers
trips17_transfers <- trips17 |>
  mutate(HOUSE_PERSONID = paste(HOUSEID, PERSONID, sep = "-")) |>
  mutate(transfer = ifelse(WHYFROM == "07" | WHYTO == "07", 1, 0)) |>
  filter(transfer == 1) |>
  select(-c("transfer"))

#identify the trips that are in a single journey, separated by transfers
trips17_transfers_chain <- trips17_transfers |>
  group_by(HOUSE_PERSONID) |>
  arrange(HOUSE_PERSONID,TDTRPNUM) |>
  mutate(chain_end = ifelse(WHYFROM == "07" & lag(WHYTO == "07") & WHYTO != "07", 1, 0)) |> #identifies the trip that is the last in the series
  mutate(chain = cumsum(chain_end == 0 & lag(chain_end, default = 0) == 1)) |> #assigns the same 'chain' numeric value to all trips in a given series
  ungroup() |>
  group_by(HOUSE_PERSONID, chain) |>
  mutate(ENDTIME = max(ENDTIME))|> #takes the endtime of the last trip in the series
  mutate(WHYFROM = first(WHYFROM)) |> #takes the origin of the first trip in the series
  mutate(WHYTO = last(WHYTO)) |> #takes the destination of the last trip in the series
  mutate(TRPTRANS = case_when(TRPMILES == max(TRPMILES) ~ TRPTRANS)) |> #takes the mode of the longest trip by distance
  mutate(TRPMILES = sum(TRPMILES)) |> #sums distance across all trips in the series
  ungroup() |>
  filter(!is.na(TRPTRANS)) |>
  select(-c("chain", "chain_end"))

#Combine non-transfer trips and transfer-trips which have now been linked into single trips
trips17b <- bind_rows(trips17_excl_transfers, trips17_transfers_chain)|>
  arrange(HOUSE_PERSONID, TDTRPNUM)|>
  select(-c("HOUSE_PERSONID"))

```

### Combine 2022 transfer trips 

```{r}

#import 2022 trips data and add a column indicating year
trips22 <- here("nhts",
                "data2022", 
                "tripv2pub.csv") |>
  read_csv(show_col_types = FALSE) |>
  mutate(year = 2022)|>
  arrange(HOUSEID, PERSONID, ENDTIME)

#generate a version of 2022 trips without transfer trips  
trips22_excl_transfers <- trips22 |>
  mutate(HOUSE_PERSONID = paste(HOUSEID, PERSONID, sep = "-")) |>
  mutate(transfer = ifelse(WHYFROM == "09" | WHYTO == "09", 1, 0)) |>
  filter(transfer == 0) |>
  select(-c("transfer"))

#generate a version of 2022 trips with only the trips involving transfers
trips22_transfers <- trips22 |>
  mutate(HOUSE_PERSONID = paste(HOUSEID, PERSONID, sep = "-")) |>
  mutate(transfer = ifelse(WHYFROM == "09" | WHYTO == "09", 1, 0)) |>
  filter(transfer == 1) |>
  select(-c("transfer"))

#identify the trips that are in a single journey, separated by transfers
trips22_transfers_chain <- trips22_transfers |>
  group_by(HOUSE_PERSONID) |>
  arrange(HOUSE_PERSONID,TRIPID) |>
  mutate(chain_end = ifelse(WHYFROM == "09" & lag(WHYTO == "09") & WHYTO != "09", 1, 0)) |> #identifies the trip that is the last in the series
  mutate(chain = cumsum(chain_end == 0 & lag(chain_end, default = 0) == 1)) |> #assigns the same 'chain' numeric value to all trips in a given series
  ungroup() |>
  group_by(HOUSE_PERSONID, chain) |>
  mutate(ENDTIME = max(ENDTIME))|> #takes the endtime of the last trip in the series
  mutate(WHYFROM = first(WHYFROM)) |> #takes the origin of the first trip in the series
  mutate(WHYTO = last(WHYTO)) |> #takes the destination of the last trip in the series
  mutate(TRPTRANS = case_when(TRPMILES == max(TRPMILES) ~ TRPTRANS)) |> #takes the mode of the longest trip by distance
  mutate(TRPMILES = sum(TRPMILES)) |> #sums distance across all trips in the series
  ungroup() |>
  filter(!is.na(TRPTRANS)) |>
  select(-c("chain", "chain_end"))

#Combine non-transfer trips and transfer-trips which have now been linked into single trips
trips22b <- bind_rows(trips22_excl_transfers, trips22_transfers_chain)|>
  arrange(HOUSE_PERSONID,TRIPID)|>
  select(-c("HOUSE_PERSONID"))

```

### Construct and combine trips data for the three years 

For each year, we filter to only include modes: walk, bike, and car 
(including van, SUV, or ride share). This excludes school bus and transit trips.
We call 'walk' or 'bike' modes "Active" modes and all other included modes "car".

For each year, we make the possible number of people on the trip the same 
(max 10) by filtering out #11+ from 2017 and 2009. 

Then, we rename the sex variables and combine the three datasets for 
2009, 2017 and 2022. For this analysis, we only consider short trips (<1.25miles) 
therefore filter accordingly. 

```{r}

#augment 2017 data
trips17b <- trips17b |>
  filter(TRPTRANS %in% c("01",
                           "02",
                           "03",
                           "04",
                           "05",
                           "06",
                           "18")) |>
  mutate(mode = ifelse(TRPTRANS %in% c("01", "02"), "Active", "Car")) |>
  rename(sex = R_SEX_IMP) |>
  select(PERSONID,
         HOUSEID,
         year,
         R_AGE,
         WHYTO,
         WHYFROM,
         ENDTIME,
         TRPMILES,
         mode,
         starts_with("ONTD_P"),
         NUMONTRP,
         NONHHCNT,
         sex,
         HHFAMINC,
         DRVRCNT,
         HHVEHCNT) 

#augment 2009 data
trips09 <- trips09 |>
  select (-ONTD_P14,
         -ONTD_P15) |> #removes columns 14 and 15 as these have no data in 2009; allows bind with 2017/22
  filter(TRPTRANS %in% c("01",
                           "02",
                           "03",
                           "04",
                           "05",
                           "22",
                           "23")) |>
  mutate(mode = ifelse(TRPTRANS %in% c("22", "23"), "Active", "Car")) |>
  rename(sex = R_SEX) |>
  select(PERSONID,
         HOUSEID,
         year,
         R_AGE,
         WHYTO,
         WHYFROM,
         ENDTIME,
         TRPMILES,
         mode,
         starts_with("ONTD_P"),
         NUMONTRP,
         NONHHCNT,
         sex,
         HHFAMINC,
         DRVRCNT,
         HHVEHCNT)

#augment 2022 data
trips22b <- trips22b |>
  mutate (ONTD_P11 = 0,
         ONTD_P12 = 0,
         ONTD_P13 = 0) |> #add columns to match 2009 and 2017 which go up to person 13
  filter(TRPTRANS %in% c("01",
                           "02",
                           "03",
                           "04",
                           "18",
                           "19",
                           "20")) |>
  mutate(mode = ifelse(TRPTRANS %in% c("18", "19","20"), "Active", "Car")) |>
  rename(sex = R_SEX_IMP) |> 
  select(PERSONID,
         HOUSEID,
         year,
         R_AGE,
         WHYTO,
         WHYFROM,
         ENDTIME,
         TRPMILES,
         mode,
         starts_with("ONTD_P"),
         NUMONTRP,
         NONHHCNT,
         sex,
         HHFAMINC,
         DRVRCNT,
         HHVEHCNT) 

#combine three files
trips <- rbind(trips09, trips17b, trips22b) |>
  mutate(arrival = as.numeric(ENDTIME)) |>
  mutate(person_hh_year = paste(PERSONID, HOUSEID, year, sep = "-"),
         hh_year = paste(HOUSEID, year, sep = "-")) |>
  group_by(person_hh_year) |>
  mutate(first_arrival = min(arrival)) |>
  ungroup() |>
  filter(R_AGE > 6 &
         R_AGE < 15 & #only keeps kids trips
         TRPMILES < 1.25) #filters longer trips
```

Check sample size by year

```{r}
table(trips$year)
```
##Household data

Next, we add the household density variable from the household data. 

We consider two density variables:
- people/sq mile (for 2022, this is approximated from the urban/rural categorization)
- urban or rural categorization

We start by loading the household-level data files for three years, selecting the
same set of variables from each year, adding a variable to indicate which
year's sample the household is from. For 2022, we add the HBPPOPDN variable and
approximate it: 
- if Urban, we assume this value is 7000
- if rural, we assume this value is 50.

Then we combine the three datasets for 2009, 2017 and 2022. 

```{r}

#import and filter 2017 household file
hh17 <- here("nhts",
                 "data2017", 
                 "hhpub.csv") |>
  read_csv(show_col_types = FALSE) |>
  select(HOUSEID,
         HBPPOPDN,
         URBRUR) |>
  mutate(year = 2017)

#import and filter 2009 household file
hh09 <- here("nhts",
               "data2009", 
               "HHV2PUB.csv") |>
  read_csv(show_col_types = FALSE) |>
  select(HOUSEID,
         HBPPOPDN,
         URBRUR) |>
  mutate(year = 2009)

#import and filter 2022 household file
hh22 <- here("nhts",
                 "data2022", 
                 "hhv2pub.csv") |>
  read_csv(show_col_types = FALSE) |>
  mutate(HBPPOPDN = ifelse(URBRUR == "01", 7000, 50)) |> #given 2022 does not have a household density variable, it is approximated from the urban/rural variable
  select(HOUSEID,
         HBPPOPDN,
         URBRUR) |>
  mutate(year = 2022)

#combine three household files
hh_density <- rbind(hh09, hh17, hh22) |>
  mutate(hh_year = paste(HOUSEID, year, sep = "-")) |>
  rename(density = HBPPOPDN,
         density_alt = URBRUR)

```

# Join household member data to trips data

```{r}
trips <- trips |>
  left_join(hh_ages) |>
  left_join(who_in_hh)
```
# Join household density data to trips data

```{r}
trips <- trips |>
  left_join(hh_density)
```
# Construct variables

We want to include the following variables:

* `purpose`: indicates the trips destination or purpose
* `school`: whether this was a school trip or not
* `log_inc_k`: Income (log-transformed, continuous)
* `veh_per_driver`: Vehicles per driver
* `non_work_mom`: Presence of non-working adult woman in household
* `non_work_dad`: Presence of non-working adult man in household
* `age`: Age
* `female`: Sex
* `has_lil_sib`: Presence of younger child in household
* `has_big_sib`: Presence of older child in household
* `log_distance`: Distance
* `log_density`: Population density of home block group
* `density_alt`: Categorical density variable (urban or rural)

And the outcome variable, `mode_ind`, which can take any of the following values:
    
* 18: Active mode accompanied by a household adult
* 28: Active mode unaccompanied by a household adult
* 17: In a car accompanied by a household adult
* 27: In a car unaccompanied by a household adult

We assume that all of the above alternatives are available for all respondents,
so we will also create the following availability variables, and set the values
to 1/TRUE for all respondents.

* av_kid_car
* av_par_car
* av_kid_act
* av_par_act

## Predictor variables

### Construct trip purpose variable 

Given each year's trip purpose is coded differently, here we consistently 
code variables across years.

```{r}
trips <- trips |>
  mutate(purpose = case_when(WHYTO == "01" & year == 2017 ~ "home",
                              WHYTO == "02" & year == 2017 ~ "home",
                              WHYTO == "03" & year == 2017 ~ "work",
                              WHYTO == "04" & year == 2017 ~ "work",
                              WHYTO == "05" & year == 2017 ~ "work",
                              WHYTO == "06" & year == 2017 ~ "pick or drop",
                              WHYTO == "07" & year == 2017 ~ "transfer",
                              WHYTO == "08" & year == 2017 ~ "school",
                              WHYTO == "09" & year == 2017 ~ "childcare",
                              WHYTO == "10" & year == 2017 ~ "adultcare",
                              WHYTO == "11" & year == 2017 ~ "errands",
                              WHYTO == "12" & year == 2017 ~ "errands",
                              WHYTO == "13" & year == 2017 ~ "errands",
                              WHYTO == "14" & year == 2017 ~ "errands",
                              WHYTO == "15" & year == 2017 ~ "recreation",
                              WHYTO == "16" & year == 2017 ~ "recreation",
                              WHYTO == "17" & year == 2017 ~ "social",
                              WHYTO == "18" & year == 2017 ~ "healthcare",
                              WHYTO == "19" & year == 2017 ~ "religion",
                              WHYTO == "01" & year == 2022 ~ "home",
                              WHYTO == "02" & year == 2022 ~ "home",
                              WHYTO == "03" & year == 2022 ~ "work",
                              WHYTO == "04" & year == 2022 ~ "work",
                              WHYTO == "05" & year == 2022 ~ "work",
                              WHYTO == "06" & year == 2022 ~ "school",
                              WHYTO == "07" & year == 2022 ~ "care",
                              WHYTO == "08" & year == 2022 ~ "work",
                              WHYTO == "09" & year == 2022 ~ "transfer",
                              WHYTO == "10" & year == 2022 ~ "pick or drop",
                              WHYTO == "11" & year == 2022 ~ "healthcare",
                              WHYTO == "12" & year == 2022 ~ "errands",
                              WHYTO == "13" & year == 2022 ~ "errands",
                              WHYTO == "14" & year == 2022 ~ "errands",
                              WHYTO == "15" & year == 2022 ~ "recreation",
                              WHYTO == "16" & year == 2022 ~ "recreation",
                              WHYTO == "17" & year == 2022 ~ "Social",
                              WHYTO == "18" & year == 2022 ~ "recreation",
                              WHYTO == "19" & year == 2022 ~ "religion",
                              WHYTO == "01" & year == 2009 ~ "home",
                              WHYTO == "10" & year == 2009 ~ "work",
                              WHYTO == "11" & year == 2009 ~ "work",
                              WHYTO == "12" & year == 2009 ~ "work",
                              WHYTO == "13" & year == 2009 ~ "work",
                              WHYTO == "14" & year == 2009 ~ "work",
                              WHYTO == "20" & year == 2009 ~ "religious",
                              WHYTO == "21" & year == 2009 ~ "school",
                              WHYTO == "22" & year == 2009 ~ "religious",
                              WHYTO == "23" & year == 2009 ~ "errands",
                              WHYTO == "24" & year == 2009 ~ "care",
                              WHYTO == "30" & year == 2009 ~ "healthcare",
                              WHYTO == "40" & year == 2009 ~ "errands",
                              WHYTO == "41" & year == 2009 ~ "errands",
                              WHYTO == "42" & year == 2009 ~ "errands",
                              WHYTO == "43" & year == 2009 ~ "errands",
                              WHYTO == "50" & year == 2009 ~ "recreation",
                              WHYTO == "51" & year == 2009 ~ "recreation",
                              WHYTO == "52" & year == 2009 ~ "recreation",
                              WHYTO == "53" & year == 2009 ~ "social",
                              WHYTO == "54" & year == 2009 ~ "recreation",
                              WHYTO == "55" & year == 2009 ~ "recreation",
                              WHYTO == "60" & year == 2009 ~ "errands",
                              WHYTO == "61" & year == 2009 ~ "errands",
                              WHYTO == "62" & year == 2009 ~ "social",
                              WHYTO == "63" & year == 2009 ~ "errands",
                              WHYTO == "64" & year == 2009 ~ "errands",
                              WHYTO == "65" & year == 2009 ~ "work",
                              WHYTO == "70" & year == 2009 ~ "transfer",
                              WHYTO == "71" & year == 2009 ~ "pick or drop",
                              WHYTO == "72" & year == 2009 ~ "pick or drop",
                              WHYTO == "73" & year == 2009 ~ "pick or drop",
                              WHYTO == "80" & year == 2009 ~ "social",
                              WHYTO == "81" & year == 2009 ~ "social",
                              WHYTO == "82" & year == 2009 ~ "social",
                              WHYTO == "83" & year == 2009 ~ "social",
                              TRUE ~ "NA"))
```

### Construct school variable 

This is a binary variable that indicates whether the trip was to school or not.

```{r}
trips <- trips |>
  mutate(schooltrip = case_when(purpose == "school" ~ 1,
                   TRUE ~ 0))
                             
```                             
                             
### Construct income variable

To construct the income variable, we first transform the values to the 
midpoint of the survey category. Then, we inflate/deflate the income to be 
equivalent to the 2017 value (multiply 2009 by 1.14; multiply 2022 by 0.86).

The top income bracket is determined by calculating the highest median income
using census data for the year.

```{r}
trips <- trips |>
  mutate(income_k = case_when(HHFAMINC == "01" & year == 2017 ~ 5,
                              HHFAMINC == "02" & year == 2017 ~ 12.5,
                              HHFAMINC == "03" & year == 2017 ~ 20,
                              HHFAMINC == "04" & year == 2017 ~ 30,
                              HHFAMINC == "05" & year == 2017 ~ 42.5,
                              HHFAMINC == "06" & year == 2017 ~ 62.5,
                              HHFAMINC == "07" & year == 2017 ~ 87.5,
                              HHFAMINC == "08" & year == 2017 ~ 112.5,
                              HHFAMINC == "09" & year == 2017 ~ 137.5,
                              HHFAMINC == "10" & year == 2017 ~ 175,
                              HHFAMINC == "11" & year == 2017 ~ 281.36,
                              HHFAMINC == "01" & year == 2009 ~ 2.5*1.14,
                              HHFAMINC == "02" & year == 2009 ~ 7.5*1.14,
                              HHFAMINC == "03" & year == 2009 ~ 12.5*1.14,
                              HHFAMINC == "04" & year == 2009 ~ 17.5*1.14,
                              HHFAMINC == "05" & year == 2009 ~ 22.5*1.14,
                              HHFAMINC == "06" & year == 2009 ~ 27.5*1.14,
                              HHFAMINC == "07" & year == 2009 ~ 32.5*1.14,
                              HHFAMINC == "08" & year == 2009 ~ 37.5*1.14,
                              HHFAMINC == "09" & year == 2009 ~ 42.5*1.14,
                              HHFAMINC == "10" & year == 2009 ~ 47.5*1.14,
                              HHFAMINC == "11" & year == 2009 ~ 52.5*1.14,
                              HHFAMINC == "12" & year == 2009 ~ 57.5*1.14,
                              HHFAMINC == "13" & year == 2009 ~ 62.5*1.14,
                              HHFAMINC == "14" & year == 2009 ~ 67.5*1.14,
                              HHFAMINC == "15" & year == 2009 ~ 72.5*1.14,
                              HHFAMINC == "16" & year == 2009 ~ 77.5*1.14,
                              HHFAMINC == "17" & year == 2009 ~ 90*1.14,
                              HHFAMINC == "18" & year == 2009 ~ 140*1.14,
                              HHFAMINC == "01" ~ 5*0.86,
                              HHFAMINC == "02" ~ 12.5*0.86,
                              HHFAMINC == "03" ~ 20*0.86,
                              HHFAMINC == "04" ~ 30*0.86,
                              HHFAMINC == "05" ~ 42.5*0.86,
                              HHFAMINC == "06" ~ 62.5*0.86,
                              HHFAMINC == "07" ~ 87.5*0.86,
                              HHFAMINC == "08" ~ 112.5*0.86,
                              HHFAMINC == "09" ~ 137.5*0.86,
                              HHFAMINC == "10" ~ 175*0.86,
                              HHFAMINC == "11" ~ 281.36*0.86,
                              TRUE ~ -9)) |>
  filter(income_k > 0) |>
  mutate(log_inc_k = log(income_k))
```

#### Check: Compare income distribution and means for 2009, 2017, 2022

```{r}

ggplot(trips[trips$year == 2009, ], aes(x = income_k)) +
  geom_histogram(binwidth = 20, fill = "grey", color = "black") +
  ggtitle("Income Distribution in 2009") +
  xlab("Income (in thousands)") +
  ylab("Frequency")

ggplot(trips[trips$year == 2017, ], aes(x = income_k)) +
  geom_histogram(binwidth = 20, fill = "grey", color = "black") +
  ggtitle("Income Distribution in 2017") +
  xlab("Income (in thousands)") +
  ylab("Frequency")

ggplot(trips[trips$year == 2022, ], aes(x = income_k)) +
  geom_histogram(binwidth = 20, fill = "grey", color = "black") +
  ggtitle("Income Distribution in 2022") +
  xlab("Income (in thousands)") +
  ylab("Frequency")

income_2009 <- trips$income_k[trips$year == 2009]
income_2017 <- trips$income_k[trips$year == 2017]
income_2022 <- trips$income_k[trips$year == 2022]

# Compute mean and standard deviation for each sample
summary_stats_income <- aggregate(income_k ~ year, data = trips, FUN = function(x) c(mean = mean(x), sd = sd(x)))

# Print the summary statistics
print.data.frame(summary_stats_income)

```

### Construct vehicles per driver variable

To determine the number of vehicles per driver, we divide the # of household vehicles by the number of drivers in each household. If there are no drivers in the household, we assume a value of 0.

```{r}
trips <- trips |>
  mutate(veh_per_driver = ifelse(DRVRCNT == 0, 0, HHVEHCNT / DRVRCNT))
```

#### Check: Compare vehicle per driver distribution and means for 2009, 2017, 2022

```{r}

ggplot(trips[trips$year == 2009, ], aes(x = veh_per_driver)) +
  geom_histogram(binwidth = 0.5, fill = "grey", color = "black") +
  ggtitle("Vehicles per Driver distribution in 2009") +
  xlab("Vehicles per Driver") +
  ylab("Frequency")

ggplot(trips[trips$year == 2017, ], aes(x = veh_per_driver)) +
  geom_histogram(binwidth = 0.5, fill = "grey", color = "black") +
  ggtitle("Vehicles per Driver distribution in 2017") +
  xlab("Vehicles per Driver") +
  ylab("Frequency")

ggplot(trips[trips$year == 2022, ], aes(x = veh_per_driver)) +
  geom_histogram(binwidth = 0.5, fill = "grey", color = "black") +
  ggtitle("Vehicles per Driver distribution in 2022") +
  xlab("Vehicles per Driver") +
  ylab("Frequency")

vehperdriv_2009 <- trips$veh_per_driver[trips$year == 2009]
vehperdriv_2017 <- trips$veh_per_driver[trips$year == 2017]
vehperdriv_2022 <- trips$veh_per_driver[trips$year == 2022]

# Compute summary statistics
summary_stats_vehperdriv <- aggregate(veh_per_driver ~ year, data = trips, FUN = function(x) c(mean = mean(x), sd = sd(x)))

# Print the summary statistics table
print.data.frame(summary_stats_vehperdriv)

```

### Rename age variable

```{r}
trips <- trips |>
  rename(age = R_AGE)
```

#### Check: Compare age distribution and means for 2009, 2017, 2022

```{r}

ggplot(trips[trips$year == 2009, ], aes(x = age)) +
  geom_histogram(binwidth = 1, fill = "grey", color = "black") +
  ggtitle("Age distribution in 2009") +
  xlab("Age") +
  ylab("Frequency")

ggplot(trips[trips$year == 2017, ], aes(x = age)) +
  geom_histogram(binwidth = 1, fill = "grey", color = "black") +
  ggtitle("Age distribution in 2017") +
  xlab("Age") +
  ylab("Frequency")

ggplot(trips[trips$year == 2022, ], aes(x = age)) +
  geom_histogram(binwidth = 1, fill = "grey", color = "black") +
  ggtitle("Age distribution in 2022") +
  xlab("Age") +
  ylab("Frequency")

age_2009 <- trips$age[trips$year == 2009]
age_2017 <- trips$age[trips$year == 2017]
age_2022 <- trips$age[trips$year == 2022]

# Compute summary statistics
summary_stats_age <- aggregate(age ~ year, data = trips, FUN = function(x) c(mean = mean(x), sd = sd(x)))

# Print the summary statistics table
print.data.frame(summary_stats_age)

```

### Construct female variable

This is derived from the sex variable where 02 = female, 01 = male

```{r}
trips <- trips |>
  mutate(female = ifelse(sex == "02", 1, 0))
```

### Construct non-working parent variables

This is derived from the non-working parent variable constructed from the people file

```{r}
trips <- trips |>
  mutate(non_work_dad = ifelse(non_work_dad, 1, 0),
         non_work_mom = ifelse(non_work_mom, 1, 0))
```

### Construct sibling variables

If an individual's age is not the same as the youngest child in the household, 
they are assumed to have a little sibling (i.e., not be the youngest child); 
if their age is not the same as the oldest child in the household, they are 
assumed to have an older sibling (i.e., they are not the oldest child).

```{r}
trips <- trips |>
  mutate(has_lil_sib = ifelse(age != youngest_kid, 1, 0),
         has_big_sib = ifelse(age != oldest_kid, 1, 0))
```

### Construct distance and density variables

First, we determine the log of continuous distance and density variables (noting the density variable for 2022 are assumed 0, so the density values will be skewed).

```{r}

trips$density <- as.numeric(trips$density) #ensures all values in density variable are numeric

trips <- trips |>
  filter(TRPMILES > 0,
         density > 0) |>
  mutate(log_distance = log(TRPMILES),
         log_density = log(density))
```

#### Check: Compare distance distribution and means for 2009, 2017, 2022

```{r}

ggplot(trips[trips$year == 2009, ], aes(x = TRPMILES)) +
  geom_histogram(binwidth = 0.05, fill = "grey", color = "black") +
  ggtitle("Distance distribution in 2009") +
  xlab("Distance") +
  ylab("Frequency")

ggplot(trips[trips$year == 2017, ], aes(x = TRPMILES)) +
  geom_histogram(binwidth = 0.05, fill = "grey", color = "black") +
  ggtitle("Distance distribution in 2017") +
  xlab("Distance") +
  ylab("Frequency")

ggplot(trips[trips$year == 2022, ], aes(x = TRPMILES)) +
  geom_histogram(binwidth = 0.05, fill = "grey", color = "black") +
  ggtitle("Distance distribution in 2022") +
  xlab("Distance") +
  ylab("Frequency")

distance_2009 <- trips$TRPMILES[trips$year == 2009]
distance_2017 <- trips$TRPMILES[trips$year == 2017]
distance_2022<- trips$TRPMILES[trips$year == 2022]

# Compute summary statistics
summary_stats_distance <- aggregate(TRPMILES ~ year, data = trips, FUN = function(x) c(mean = mean(x), sd = sd(x)))

# Print the summary statistics table
print.data.frame(summary_stats_distance)

```

#### Check: Compare density distribution and means for 2009, 2017, 2022

```{r}

ggplot(trips[trips$year == 2009, ], aes(x = density)) +
  geom_histogram(binwidth = 1000, fill = "grey", color = "black") +
  ggtitle("Density distribution in 2009") +
  xlab("Density") +
  ylab("Frequency")

ggplot(trips[trips$year == 2017, ], aes(x = density)) +
  geom_histogram(binwidth = 1000, fill = "grey", color = "black") +
  ggtitle("Density distribution in 2017") +
  xlab("Density") +
  ylab("Frequency")

ggplot(trips[trips$year == 2022, ], aes(x = density)) +
  geom_histogram(binwidth = 1000, fill = "grey", color = "black") +
  ggtitle("Density distribution in 2022") +
  xlab("Density") +
  ylab("Frequency")

density_2009 <- trips$density[trips$year == 2009]
density_2017 <- trips$density[trips$year == 2017]
density_2022<- trips$density[trips$year == 2022]

# Compute summary statistics
summary_stats_density <- aggregate(density ~ year, data = trips, FUN = function(x) c(mean = mean(x), sd = sd(x)))

# Print the summary statistics table
print.data.frame(summary_stats_density)

```

### Construct year variable

```{r}
trips <- trips |>
  mutate(y2017 = ifelse(year == 2017, 1, 0)) |>
  mutate(y2009 = ifelse(year == 2009, 1, 0))
```

#### Check: Check sample size by year

```{r}
table(trips$year)
```

## Outcome variable

First, we add a variable to indicate whether the trip is accompanied by an adult 
(assumed parent) or not. This is the case if an adult (age>17) is on the trip.

Then, we construct the mode_ind (mode and independence status) for the trip 
according to the following rules:
* 18: Active mode accompanied by a household adult
* 28: Active mode unaccompanied by a household adult
* 17: In a car accompanied by a household adult
* 27: In a car unaccompanied by a household adult

```{r}
trips <- trips |>
  mutate(with_parent = (ONTD_P1 == "01" & age_01 > 17) |
           (ONTD_P2 == "01" & age_02 > 17) |
           (ONTD_P3 == "01" & age_03 > 17) |
           (ONTD_P4 == "01" & age_04 > 17) |
           (ONTD_P5 == "01" & age_05 > 17) |
           (ONTD_P6 == "01" & age_06 > 17) |
           (ONTD_P7 == "01" & age_07 > 17) |
           (ONTD_P8 == "01" & age_08 > 17) |
           (ONTD_P9 == "01" & age_09 > 17) |
           (ONTD_P10 == "01" & age_10 > 17)|
           (ONTD_P11 == "01" & age_11 > 17) |
           (ONTD_P12 == "01" & age_12 > 17) |
           (ONTD_P13 == "01" & age_13 > 17)) |>
  mutate(mode_ind = case_when(with_parent & mode == "Active" ~ 18,
                              with_parent & mode == "Car" ~ 17,
                              mode == "Active" ~ 28, #not with parent
                              mode == "Car" ~ 27, #not with parent
                              TRUE ~ -1))
```

## Availability variables

We assume that all of the above alternatives are available for all respondents,
so we will also create the following availability variables, and set the values
to 1/TRUE for all respondents.

```{r}
trips <- trips |>
  mutate(av_par_car = 1,
         av_par_act = 1,
         av_kid_car = 1,
         av_kid_act = 1)
```

# Select final variable set

```{r}
trips <- trips |>
  select(mode_ind,
         schooltrip,
         y2017,
         av_par_car,
         av_par_act,
         av_kid_car,
         av_kid_act,
         log_inc_k,
         veh_per_driver,
         non_work_dad,
         non_work_mom,
         age,
         female,
         has_lil_sib,
         has_big_sib,
         log_distance,
         log_density,
         density_alt)
  
```

# Write csv file

```{r}
write_csv(trips,
          here("models",
               "IATBR plan",
               "trips.csv"))
```

# scale variables 1

```{r}
trips_scaled <- trips |>
  mutate(sc_y2017 = y2017/100,
         sc_log_inc_k = log_inc_k/100,
         sc_age = age/100,
         sc_veh_per_driver = veh_per_driver/100,
         sc_non_work_dad = non_work_dad/100,
         sc_non_work_mom = non_work_mom/100,
         sc_female = female/100,
         sc_has_lil_sib = has_lil_sib/100,
         sc_has_big_sib = has_big_sib/100,
         sc_log_distance = log_distance/100,
         sc_log_density = log_density/100) |>
  select(mode_ind, 
         sc_y2017,
         av_par_car,
         av_par_act,
         av_kid_car,
         av_kid_act,
         sc_log_inc_k,
         sc_veh_per_driver,
         sc_non_work_dad,
         sc_non_work_mom,
         sc_age,
         sc_female,
         sc_has_lil_sib,
         sc_has_big_sib,
         sc_log_distance,
         sc_log_density)
```

# Write csv file

```{r}
write_csv(trips_scaled,
          here("models",
               "IATBR plan",
               "trips_sc.csv"))
```

# scale variables 2

```{r}
trips_scaled2 <- trips |>
  mutate(sc2_y2017 = y2017/10,
         sc2_log_inc_k = log_inc_k/10,
         sc2_age = age/10,
         sc2_veh_per_driver = veh_per_driver/100,
         sc2_non_work_dad = non_work_dad/10,
         sc2_non_work_mom = non_work_mom/10,
         sc2_female = female/10,
         sc2_has_lil_sib = has_lil_sib/10,
         sc2_has_big_sib = has_big_sib/10,
         sc2_log_density = log_density/10) |>
  select(mode_ind, 
         sc2_y2017,
         av_par_car,
         av_par_act,
         av_kid_car,
         av_kid_act,
         sc2_log_inc_k,
         sc2_veh_per_driver,
         sc2_non_work_dad,
         sc2_non_work_mom,
         sc2_age,
         sc2_female,
         sc2_has_lil_sib,
         sc2_has_big_sib,
         log_distance,
         sc2_log_density)
```

# Write csv file

```{r}
write_csv(trips_scaled2,
          here("models",
               "IATBR plan",
               "trips_sc2.csv"))
```

# Create one with combined car alternatives

Here, we combine the two mode_ind involving car travel into a single outcome 
variable (7) because we assume most car trips are accompanied. This results in 
three outcome variables:
* 18: Active mode accompanied by a household adult
* 28: Active mode unaccompanied by a household adult
* 7: In a car (accompanied or unaccompanied)

```{r}
trips3 <- trips |>
  mutate(mode_ind = ifelse(mode_ind %in% c(17, 27), 7, mode_ind ))

write_csv(trips3,
          here("models",
               "IATBR plan",
               "trips3.csv"))
```

# Summaries

```{r}
trips |>
  group_by(y2017) |>
  summarise(car_with_parent = sum(mode_ind == 17),
            car_no_parent = sum(mode_ind == 27),
            active_with_parent = sum(mode_ind == 18),
            active_no_parent = sum(mode_ind == 28))
```


```{r}
trips |>
  group_by(y2017) |>
  summarise(car_with_parent = sum(mode_ind == 17)/n(),
            car_no_parent = sum(mode_ind == 27)/n(),
            active_with_parent = sum(mode_ind == 18)/n(),
            active_no_parent = sum(mode_ind == 28)/n())
```

```{r}
trips |>
  summarise(car_with_parent = sum(mode_ind == 17),
            car_no_parent = sum(mode_ind == 27),
            active_with_parent = sum(mode_ind == 18),
            active_no_parent = sum(mode_ind == 28))
```


```{r}
trips |>
  summarise(car_with_parent = sum(mode_ind == 17)/n(),
            car_no_parent = sum(mode_ind == 27)/n(),
            active_with_parent = sum(mode_ind == 18)/n(),
            active_no_parent = sum(mode_ind == 28)/n())
```

