---
title: "IATBR data assembly"
output: html_document
date: "2024-04-01"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Required libraries

We're only using the `tidyverse` and `here` libraries.

```{r, message=FALSE}

library(tidyverse)
library(here)
```

# Assemble person-level data

We need information on the ages of the household members on each trip to 
determine whether a child was accompanied by a household adult.

We also need to determine whether each household includes a non-working female
adults and/or a non-working male adult and the ages of the youngest and 
oldest children in each household (to determine if the trip-maker has a 
younger or older sibling).

## Load person-level data

We start by loading the person-level data files for both years, selecting the
same set of variables from each year, adding a variable to indicate which
year's sample the household is from, and renaming the sex variable to have the
same name in both years.

Then we combine the two datasets. 

```{r}
people17 <- here("nhts",
                 "data2017", 
                 "perpub.csv") |>
  read_csv(show_col_types = FALSE) |>
  select(HOUSEID, 
         PERSONID,
         R_AGE,
         R_SEX_IMP,
         WORKER,
         HHSIZE) |>
  mutate(year = 2017) |>
  rename(sex = R_SEX_IMP)

people09 <- here("nhts",
               "data2009", 
               "PERV2PUB.csv") |>
  read_csv(show_col_types = FALSE) |>
  select(HOUSEID, 
         PERSONID,
         R_AGE,
         R_SEX,
         WORKER,
         HHSIZE) |>
  mutate(year = 2009) |>
  rename(sex = R_SEX)

people <- rbind(people09, people17) |>
  mutate(person_hh_year = paste(PERSONID, HOUSEID, year, sep = "-")) |>
  mutate(hh_year = paste(HOUSEID, year, sep = "-"))

```

## Get ages of household members

We'll reshape the person-level data into a household-level dataset with
a column indicating the age of each person in the household.

```{r}
hh_ages <- people |>
  select(hh_year, PERSONID, R_AGE) |>
  pivot_wider(names_from = PERSONID,
              names_prefix = "age_",
              values_from = R_AGE)
```

## Get presence of non-working adult men and women and age ranges of children

```{r}
who_in_hh <- people |>
  mutate(adult = R_AGE > 17,
         kid_age = ifelse(R_AGE < 18, R_AGE, -2),
         R_AGE = ifelse(R_AGE < 0, 999, R_AGE),
         non_worker_mom = R_AGE > 17 & WORKER != "01" & sex == "02",
         non_worker_dad = R_AGE > 17 & WORKER != "01" & sex != "02") |>
  group_by(hh_year, HHSIZE) |>
  summarize(num_records = n(),
         n_adults = sum(adult),
         non_work_mom = sum(non_worker_mom) > 0,
         non_work_dad = sum(non_worker_dad) > 0,
         min_age = min(R_AGE),
         max_kid_age = max(kid_age)) |>
  mutate(youngest_kid = ifelse(num_records < HHSIZE, 1, min_age),
         oldest_kid = ifelse(max_kid_age > 0, max_kid_age, 1),
         n_children = HHSIZE - n_adults) |>
  ungroup() |>
  filter(n_children > 0) |>
  select(hh_year, 
         non_work_dad,
         non_work_mom,
         oldest_kid, 
         youngest_kid)
```

## Trips data


```{r, fig.width = 4, fig.height = 4}
trips17_all_dist_mode <- here("nhts",
                "data2017", 
                "trippub.csv") |>
  read_csv(show_col_types = FALSE) |>
  mutate(year = 2017) |>
  filter(WHYFROM %in% c("01","02") & 
           WHYTO == "08") |>
  mutate(mode = case_when(TRPTRANS == "01" ~ "Walk",
                          TRPTRANS == "02" ~ "Bike",
                          TRPTRANS %in% c("03", "04", "05", "06") ~ "Car",
                          TRPTRANS == "10" ~ "School bus",
                          TRPTRANS %in% c("11", "12", "13", "14", "15", "16") ~ "Transit",
                          TRUE ~ "Other")) |>
  rename(sex = R_SEX_IMP,
         density = OBPPOPDN) |>
  select(PERSONID,
         HOUSEID,
         year,
         R_AGE,
         ENDTIME,
         TRPMILES,
         mode,
         starts_with("ONTD_P"),
         NUMONTRP,
         NONHHCNT,
         sex,
         HHFAMINC,
         DRVRCNT,
         HHVEHCNT,
         density,
         TRPTRANS) 

trips09_all_dist_mode <- here("nhts",
              "data2009", 
              "DAYV2PUB.csv") |>
  read_csv(show_col_types = FALSE) |>
  mutate(year = 2009) |>
  select(-ONTD_P15, -ONTD_P14) |>
  filter(WHYFROM == "01" & 
           WHYTO == "21") |>
  mutate(mode = case_when(TRPTRANS == "22" ~ "Bike",
                          TRPTRANS == "23" ~ "Walk", 
                          TRPTRANS %in% c("01", "02", "03", "04", "05")~ "Car",
                          TRPTRANS == "11" ~ "School bus",
                          TRPTRANS %in% c("09", "10", "12", "13", "14", 
                                          "15", "16", "17", "18") ~ "Transit",
                          TRUE ~ "Other")) |>
  rename(sex = R_SEX,
         density = HBPPOPDN) |>
  select(PERSONID,
         HOUSEID,
         year,
         R_AGE,
         ENDTIME,
         TRPMILES,
         mode,
         starts_with("ONTD_P"),
         NUMONTRP,
         NONHHCNT,
         sex,
         HHFAMINC,
         DRVRCNT,
         HHVEHCNT,
         density,
         TRPTRANS)

trips_all_dist_mode <- rbind(trips09_all_dist_mode, trips17_all_dist_mode) |>
  mutate(arrival = as.numeric(ENDTIME)) |>
  mutate(person_hh_year = paste(PERSONID, HOUSEID, year, sep = "-"),
         hh_year = paste(HOUSEID, year, sep = "-")) |>
  group_by(person_hh_year) |>
  mutate(first_arrival = min(arrival)) |>
  ungroup() |>
  filter(arrival == first_arrival,
         R_AGE > 6 &
         R_AGE < 15 & 
         arrival < 1000) |>
  mutate(short = TRPMILES < 1.25) |>
  mutate(mode = factor(mode, 
                       levels = c("Other",
                                  "Transit",
                                  "School bus",
                                  "Car",
                                  "Bike",
                                  "Walk")))

ggplot(trips_all_dist_mode) +
  geom_bar(aes(x = short, fill = mode), color = "gray") +
  scale_x_discrete(name = "Trip distance",
                   labels = c("1.25 miles\nor more",
                              "Less than\n1.25 miles")) +
  scale_y_continuous(name = "Number of respondents") +
  scale_fill_manual(values = c("gray30",
                                "gray80",
                                "white",
                                "yellow",
                                "red",
                                "blue")) +
  theme_bw()
```


```{r}
trips17 <- here("nhts",
                "data2017", 
                "trippub.csv") |>
  read_csv(show_col_types = FALSE) |>
  mutate(year = 2017) |>
  filter(WHYFROM %in% c("01","02") & 
           WHYTO == "08" &
           TRPTRANS %in% c("01",
                           "02",
                           "03",
                           "04",
                           "05",
                           "06")) |>
  mutate(mode = ifelse(TRPTRANS %in% c("01", "02"), "Active", "Car")) |>
  rename(sex = R_SEX_IMP,
         density = OBPPOPDN) |>
  select(PERSONID,
         HOUSEID,
         year,
         R_AGE,
         ENDTIME,
         TRPMILES,
         mode,
         starts_with("ONTD_P"),
         NUMONTRP,
         NONHHCNT,
         sex,
         HHFAMINC,
         DRVRCNT,
         HHVEHCNT,
         density) 

trips09 <- here("nhts",
              "data2009", 
              "DAYV2PUB.csv") |>
  read_csv(show_col_types = FALSE) |>
  mutate(year = 2009) |>
  select(-ONTD_P15, -ONTD_P14) |>
  filter(WHYFROM == "01" & 
           WHYTO == "21" &
           TRPTRANS %in% c("01",
                           "02",
                           "03",
                           "04",
                           "05",
                           "22",
                           "23")) |>
  mutate(mode = ifelse(TRPTRANS %in% c("22", "23"), "Active", "Car")) |>
  rename(sex = R_SEX,
         density = HBPPOPDN) |>
  select(PERSONID,
         HOUSEID,
         year,
         R_AGE,
         ENDTIME,
         TRPMILES,
         mode,
         starts_with("ONTD_P"),
         NUMONTRP,
         NONHHCNT,
         sex,
         HHFAMINC,
         DRVRCNT,
         HHVEHCNT,
         density)

trips <- rbind(trips09, trips17) |>
  mutate(arrival = as.numeric(ENDTIME)) |>
  mutate(person_hh_year = paste(PERSONID, HOUSEID, year, sep = "-"),
         hh_year = paste(HOUSEID, year, sep = "-")) |>
  group_by(person_hh_year) |>
  mutate(first_arrival = min(arrival)) |>
  ungroup() |>
  filter(arrival == first_arrival,
         R_AGE > 6 &
         R_AGE < 15 & 
         arrival < 1000,
         TRPMILES < 1.25) 
```



Check sample size by year

```{r}
table(trips$year)
```

# Add household member data to trips data

```{r}
trips <- trips |>
  left_join(hh_ages) |>
  left_join(who_in_hh)
```

# Construct variables

We want to include the following variables:

* `log_inc_k`: Income (log-transformed, continuous)
* `veh_per_driver` :Vehicles per driver
* `non_work_mom`: Presence of non-working adult woman in household
* `non_work_dad`: Presence of non-working adult man in household
* `age`: Age
* `female`: Sex
* `has_lil_sib`: Presence of younger child in household
* `has_big_sib`: Presence of older child in household
* `log_distance`: Distance
* `log_density`: Population density of home block group

And the outcome variable, `mode_ind`, which can take any of the following values:
    
* 18: Active mode accompanied by a household adult
* 28: Active mode unaccompanied by a household adult
* 17: In a car accompanied by a household adult
* 27: In a car unaccompanied by a household adult

We assume that all of the above alternatives are available for all respondents,
so we will also create the following availability variables, and set the values
to 1/TRUE for all respondents.

* av_kid_car
* av_par_car
* av_kid_act
* av_par_act

## Predictor variables

### Construct income variable

```{r}
trips <- trips |>
  mutate(income_k = case_when(HHFAMINC == "01" & year == 2017 ~ 5,
                              HHFAMINC == "02" & year == 2017 ~ 12.5,
                              HHFAMINC == "03" & year == 2017 ~ 20,
                              HHFAMINC == "04" & year == 2017 ~ 30,
                              HHFAMINC == "05" & year == 2017 ~ 42.5,
                              HHFAMINC == "06" & year == 2017 ~ 62.5,
                              HHFAMINC == "07" & year == 2017 ~ 87.5,
                              HHFAMINC == "08" & year == 2017 ~ 112.5,
                              HHFAMINC == "09" & year == 2017 ~ 137.5,
                              HHFAMINC == "10" & year == 2017 ~ 175,
                              HHFAMINC == "11" & year == 2017 ~ 281.36,
                              HHFAMINC == "01" ~ 2.5*1.14,
                              HHFAMINC == "02" ~ 7.5*1.14,
                              HHFAMINC == "03" ~ 12.5*1.14,
                              HHFAMINC == "04" ~ 17.5*1.14,
                              HHFAMINC == "05" ~ 22.5*1.14,
                              HHFAMINC == "06" ~ 27.5*1.14,
                              HHFAMINC == "07" ~ 32.5*1.14,
                              HHFAMINC == "08" ~ 37.5*1.14,
                              HHFAMINC == "09" ~ 42.5*1.14,
                              HHFAMINC == "10" ~ 47.5*1.14,
                              HHFAMINC == "11" ~ 52.5*1.14,
                              HHFAMINC == "12" ~ 57.5*1.14,
                              HHFAMINC == "13" ~ 62.5*1.14,
                              HHFAMINC == "14" ~ 67.5*1.14,
                              HHFAMINC == "15" ~ 72.5*1.14,
                              HHFAMINC == "16" ~ 77.5*1.14,
                              HHFAMINC == "17" ~ 90*1.14,
                              HHFAMINC == "18" ~ 140*1.14,
                              TRUE ~ -9)) |>
  filter(income_k > 0) |>
  mutate(log_inc_k = log(income_k))
```

### Compare income distribution and means for 2009 and 2017

```{r}

library(ggplot2)

ggplot(trips[trips$year == 2009, ], aes(x = income_k)) +
  geom_histogram(binwidth = 20, fill = "grey", color = "black") +
  ggtitle("Income Distribution in 2009") +
  xlab("Income (in thousands)") +
  ylab("Frequency")

ggplot(trips[trips$year == 2017, ], aes(x = income_k)) +
  geom_histogram(binwidth = 20, fill = "grey", color = "black") +
  ggtitle("Income Distribution in 2017") +
  xlab("Income (in thousands)") +
  ylab("Frequency")

income_2009 <- trips$income_k[trips$year == 2009]
income_2017 <- trips$income_k[trips$year == 2017]

# Perform t-test
t_test_result <- t.test(income_2009, income_2017)

# Print the results
print(t_test_result)

```

### Construct vehicles per driver variable

```{r}
trips <- trips |>
  mutate(veh_per_driver = ifelse(DRVRCNT == 0, 0, HHVEHCNT / DRVRCNT))
```

### Compare vehicle per driver distribution and means for 2009 and 2017

```{r}

library(ggplot2)

ggplot(trips[trips$year == 2009, ], aes(x = veh_per_driver)) +
  geom_histogram(binwidth = 0.5, fill = "grey", color = "black") +
  ggtitle("Vehicles per Driver distribution in 2009") +
  xlab("Vehicles per Driver") +
  ylab("Frequency")

ggplot(trips[trips$year == 2017, ], aes(x = veh_per_driver)) +
  geom_histogram(binwidth = 0.5, fill = "grey", color = "black") +
  ggtitle("Vehicles per Driver distribution in 2017") +
  xlab("Vehicles per Driver") +
  ylab("Frequency")

vehperdriv_2009 <- trips$veh_per_driver[trips$year == 2009]
vehperdriv_2017 <- trips$veh_per_driver[trips$year == 2017]

# Perform t-test
t_test_result <- t.test(vehperdriv_2009, vehperdriv_2017)

# Print the results
print(t_test_result)

```

### Rename age variable

```{r}
trips <- trips |>
  rename(age = R_AGE)
```

### Compare age distribution and means for 2009 and 2017

```{r}

library(ggplot2)

ggplot(trips[trips$year == 2009, ], aes(x = age)) +
  geom_histogram(binwidth = 1, fill = "grey", color = "black") +
  ggtitle("Age distribution in 2009") +
  xlab("Age") +
  ylab("Frequency")

ggplot(trips[trips$year == 2017, ], aes(x = age)) +
  geom_histogram(binwidth = 1, fill = "grey", color = "black") +
  ggtitle("Age distribution in 2017") +
  xlab("Age") +
  ylab("Frequency")

age_2009 <- trips$age[trips$year == 2009]
age_2017 <- trips$age[trips$year == 2017]

# Perform t-test
t_test_result <- t.test(age_2009, age_2017)

# Print the results
print(t_test_result)

```

### Construct female variable

```{r}
trips <- trips |>
  mutate(female = ifelse(sex == "02", 1, 0))
```

### Construct non-working parent variables

```{r}
trips <- trips |>
  mutate(non_work_dad = ifelse(non_work_dad, 1, 0),
         non_work_mom = ifelse(non_work_mom, 1, 0))
```

### Construct sibling variables

```{r}
trips <- trips |>
  mutate(has_lil_sib = ifelse(age != youngest_kid, 1, 0),
         has_big_sib = ifelse(age != oldest_kid, 1, 0))
```

### Construct distance and density variables

```{r}
trips <- trips |>
  filter(TRPMILES > 0,
         density > 0) |>
  mutate(log_distance = log(TRPMILES),
         log_density = log(density))
```

### Compare distance distribution and means for 2009 and 2017

```{r}

library(ggplot2)

ggplot(trips[trips$year == 2009, ], aes(x = TRPMILES)) +
  geom_histogram(binwidth = 0.05, fill = "grey", color = "black") +
  ggtitle("Distance distribution in 2009") +
  xlab("Distance") +
  ylab("Frequency")

ggplot(trips[trips$year == 2017, ], aes(x = TRPMILES)) +
  geom_histogram(binwidth = 0.05, fill = "grey", color = "black") +
  ggtitle("Distance distribution in 2017") +
  xlab("Distance") +
  ylab("Frequency")

distance_2009 <- trips$TRPMILES[trips$year == 2009]
distance_2017 <- trips$TRPMILES[trips$year == 2017]

# Perform t-test
t_test_result <- t.test(distance_2009, distance_2017)

# Print the results
print(t_test_result)

```

### Compare density distribution and means for 2009 and 2017

```{r}

library(ggplot2)

ggplot(trips[trips$year == 2009, ], aes(x = density)) +
  geom_histogram(binwidth = 500, fill = "grey", color = "black") +
  ggtitle("Density distribution in 2009") +
  xlab("Density") +
  ylab("Frequency")

ggplot(trips[trips$year == 2017, ], aes(x = density)) +
  geom_histogram(binwidth = 500, fill = "grey", color = "black") +
  ggtitle("Density distribution in 2017") +
  xlab("Density") +
  ylab("Frequency")

density_2009 <- trips$density[trips$year == 2009]
density_2017 <- trips$density[trips$year == 2017]

# Perform t-test
t_test_result <- t.test(density_2009, density_2017)

# Print the results
print(t_test_result)

```

### Construct year variable

```{r}
trips <- trips |>
  mutate(y2017 = ifelse(year == 2017, 1, 0))
```


### Check sample size by year

```{r}
table(trips$year)
```
## Outcome variable

```{r}
trips <- trips |>
  mutate(with_parent = (ONTD_P1 == "01" & age_01 > 17) |
           (ONTD_P2 == "01" & age_02 > 17) |
           (ONTD_P3 == "01" & age_03 > 17) |
           (ONTD_P4 == "01" & age_04 > 17) |
           (ONTD_P5 == "01" & age_05 > 17) |
           (ONTD_P6 == "01" & age_06 > 17) |
           (ONTD_P7 == "01" & age_07 > 17) |
           (ONTD_P8 == "01" & age_08 > 17) |
           (ONTD_P9 == "01" & age_09 > 17) |
           (ONTD_P10 == "01" & age_10 > 17) |
           (ONTD_P11 == "01" & age_11 > 17) |
           (ONTD_P12 == "01" & age_12 > 17) |
           (ONTD_P13 == "01" & age_13 > 17)) |>
  mutate(mode_ind = case_when(with_parent & mode == "Active" ~ 18,
                              with_parent & mode == "Car" ~ 17,
                              mode == "Active" ~ 28,
                              mode == "Car" ~ 27,
                              TRUE ~ -1))
```

## Availability variables

```{r}
trips <- trips |>
  mutate(av_par_car = 1,
         av_par_act = 1,
         av_kid_car = 1,
         av_kid_act = 1)
```

# Select final variable set

```{r}
trips <- trips |>
  select(mode_ind, 
         y2017,
         av_par_car,
         av_par_act,
         av_kid_car,
         av_kid_act,
         log_inc_k,
         veh_per_driver,
         non_work_dad,
         non_work_mom,
         age,
         female,
         has_lil_sib,
         has_big_sib,
         log_distance,
         log_density)
  
```

# Write csv file

```{r}
write_csv(trips,
          here("models",
               "IATBR plan",
               "trips.csv"))
```

# scale variables 1

```{r}
trips_scaled <- trips |>
  mutate(sc_y2017 = y2017/100,
         sc_log_inc_k = log_inc_k/100,
         sc_age = age/100,
         sc_veh_per_driver = veh_per_driver/100,
         sc_non_work_dad = non_work_dad/100,
         sc_non_work_mom = non_work_mom/100,
         sc_female = female/100,
         sc_has_lil_sib = has_lil_sib/100,
         sc_has_big_sib = has_big_sib/100,
         sc_log_distance = log_distance/100,
         sc_log_density = log_density/100) |>
  select(mode_ind, 
         sc_y2017,
         av_par_car,
         av_par_act,
         av_kid_car,
         av_kid_act,
         sc_log_inc_k,
         sc_veh_per_driver,
         sc_non_work_dad,
         sc_non_work_mom,
         sc_age,
         sc_female,
         sc_has_lil_sib,
         sc_has_big_sib,
         sc_log_distance,
         sc_log_density)
```

# Write csv file

```{r}
write_csv(trips_scaled,
          here("models",
               "IATBR plan",
               "trips_sc.csv"))
```

# scale variables 2

```{r}
trips_scaled2 <- trips |>
  mutate(sc2_y2017 = y2017/10,
         sc2_log_inc_k = log_inc_k/10,
         sc2_age = age/10,
         sc2_veh_per_driver = veh_per_driver/100,
         sc2_non_work_dad = non_work_dad/10,
         sc2_non_work_mom = non_work_mom/10,
         sc2_female = female/10,
         sc2_has_lil_sib = has_lil_sib/10,
         sc2_has_big_sib = has_big_sib/10,
         sc2_log_density = log_density/10) |>
  select(mode_ind, 
         sc2_y2017,
         av_par_car,
         av_par_act,
         av_kid_car,
         av_kid_act,
         sc2_log_inc_k,
         sc2_veh_per_driver,
         sc2_non_work_dad,
         sc2_non_work_mom,
         sc2_age,
         sc2_female,
         sc2_has_lil_sib,
         sc2_has_big_sib,
         log_distance,
         sc2_log_density)
```

# Write csv file

```{r}
write_csv(trips_scaled2,
          here("models",
               "IATBR plan",
               "trips_sc2.csv"))
```

# Create one with combined car alternatives

```{r}
trips3 <- trips |>
  mutate(mode_ind = ifelse(mode_ind %in% c(17, 27), 7, mode_ind ))

write_csv(trips3,
          here("models",
               "IATBR plan",
               "trips3.csv"))
```



# Summaries

```{r}
trips |>
  group_by(y2017) |>
  summarise(car_with_parent = sum(mode_ind == 17),
            car_no_parent = sum(mode_ind == 27),
            active_with_parent = sum(mode_ind == 18),
            active_no_parent = sum(mode_ind == 28))
```


```{r}
trips |>
  group_by(y2017) |>
  summarise(car_with_parent = sum(mode_ind == 17)/n(),
            car_no_parent = sum(mode_ind == 27)/n(),
            active_with_parent = sum(mode_ind == 18)/n(),
            active_no_parent = sum(mode_ind == 28)/n())
```

```{r}
trips |>
  summarise(car_with_parent = sum(mode_ind == 17),
            car_no_parent = sum(mode_ind == 27),
            active_with_parent = sum(mode_ind == 18),
            active_no_parent = sum(mode_ind == 28))
```


```{r}
trips |>
  summarise(car_with_parent = sum(mode_ind == 17)/n(),
            car_no_parent = sum(mode_ind == 27)/n(),
            active_with_parent = sum(mode_ind == 18)/n(),
            active_no_parent = sum(mode_ind == 28)/n())
```
