---
title: "School Trips_2009-2017-2022"
author: "Aanchal Chopra"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Required libraries

We're only using the `tidyverse`,  `here`, and `ggplot2` libraries.

```{r, message=FALSE}

library(tidyverse)
library(here)
library(ggplot2)
```

# Assemble person-level data

We need information on the ages of the household members on each trip to 
determine whether a child was accompanied by a household adult.

We also need to determine whether each household includes a non-working female
adults and/or a non-working male adult and the ages of the youngest and 
oldest children in each household (to determine if the trip-maker has a 
younger or older sibling).

## Load person-level data

We start by loading the person-level data files for three years, selecting the
same set of variables from each year, adding a variable to indicate which
year's sample the household is from, and renaming the sex variable to have the
same name in all three years.

Then we combine the three datasets for 2009, 2017 and 2022. 

```{r}

#import and filter 2017 person file
people17 <- here("nhts",
                 "data2017", 
                 "perpub.csv") |>
  read_csv(show_col_types = FALSE) |>
  select(HOUSEID, 
         PERSONID,
         R_AGE,
         R_SEX_IMP,
         WORKER,
         HHSIZE) |>
  mutate(year = 2017) |>
  rename(sex = R_SEX_IMP)

#import and filter 2009 person file
people09 <- here("nhts",
               "data2009", 
               "PERV2PUB.csv") |>
  read_csv(show_col_types = FALSE) |>
  select(HOUSEID, 
         PERSONID,
         R_AGE,
         R_SEX,
         WORKER,
         HHSIZE) |>
  mutate(year = 2009) |>
  rename(sex = R_SEX)

#import and filter 2022 person file
people22 <- here("nhts",
                 "data2022", 
                 "perv2pub.csv") |>
  read_csv(show_col_types = FALSE) |>
  select(HOUSEID, 
         PERSONID,
         R_AGE,
         R_SEX_IMP,
         WORKER,
         HHSIZE) |>
  mutate(year = 2022) |>
  rename(sex = R_SEX_IMP)

#combine three person files
people <- rbind(people09, people17, people22) |>
  mutate(person_hh_year = paste(PERSONID, HOUSEID, year, sep = "-")) |>
  mutate(hh_year = paste(HOUSEID, year, sep = "-"))

```

## Get ages of household members

We'll reshape the person-level data into a household-level dataset with
a column indicating the age of each person in the household.

```{r}
hh_ages <- people |>
  select(hh_year, PERSONID, R_AGE) |>
  pivot_wider(names_from = PERSONID,
              names_prefix = "age_",
              values_from = R_AGE)
```

## Get presence of non-working adult men and women and age ranges of children

For each household, we'll first determine which households include an adult 
(if age is greater than 17 i.e., 18 or above) or kid (age < 18 i.e., 17 or 
below). For adults, we assume all female adults are mothers and all male adults 
are fathers (or perform equivalent supervisory functions). 

Then, we determine which parents are working. We'll then reshape this 
household-level data to determine the presence of non-working adult men and 
women and the age of the oldest and youngest children.

```{r}
who_in_hh <- people |>
  mutate(adult = R_AGE > 17,
         kid_age = ifelse(R_AGE < 18, R_AGE, -2),
         R_AGE = ifelse(R_AGE < 0, 999, R_AGE),
         non_worker_mom = R_AGE > 17 & WORKER != "01" & sex == "02", #non-working parent is assumed to be an adult (>17) and not a Worker (01)
         non_worker_dad = R_AGE > 17 & WORKER != "01" & sex != "02") |>
  group_by(hh_year, HHSIZE) |>
  summarize(num_records = n(),
         n_adults = sum(adult),
         non_work_mom = sum(non_worker_mom) > 0,
         non_work_dad = sum(non_worker_dad) > 0,
         min_age = min(R_AGE),
         max_kid_age = max(kid_age)) |>
  mutate(youngest_kid = ifelse(num_records < HHSIZE, 1, min_age),
         oldest_kid = ifelse(max_kid_age > 0, max_kid_age, 1),
         n_children = HHSIZE - n_adults) |>
  ungroup() |>
  filter(n_children > 0) |>
  select(hh_year, 
         non_work_dad,
         non_work_mom,
         oldest_kid, 
         youngest_kid)
```
## Trips data

We then load in the trip-level data files for the three years, adding a variable
to indicate the year of the sample.

For each year, we filter the data set to only include trips in which the 
'destination is 'school' and the origin is 'home'. We also filter to only 
include modes: walk, bike, and car (including van, SUV, or ride share). 
This excludes school bus and transit trips.We call 'walk' or 'bike' modes 
"Active" modes and all other included modes "car".

For each year, we make the possible number of people on the trip the same 
(max 10) by filtering out #11+ from 2017 and 2009. 

An alternative density variable is also created (density_alt) to align with the 
available density variables for 2022; this is a categorical variable. 

Then, we rename the sex and density variables and combine the three datasets for 
2009, 2017 and 2022. 

Finally, we filter to capture the first trip of the day which includes a child 
only and is prior to 10am (assuming this is a schooltrip).For this analysis, we 
only consider short trips (<1.25miles) therefore filter accordingly. 

```{r}

#import and filter 2017 trips data
trips17 <- here("nhts",
                "data2017", 
                "trippub.csv") |>
  read_csv(show_col_types = FALSE) |>
  mutate(year = 2017) |>
  filter(WHYFROM %in% c("01","02") & 
           WHYTO == "08" &
           TRPTRANS %in% c("01",
                           "02",
                           "03",
                           "04",
                           "05",
                           "06",
                           "18")) |>
  mutate(mode = ifelse(TRPTRANS %in% c("01", "02"), "Active", "Car")) |>
  rename(sex = R_SEX_IMP,
         density = OBPPOPDN,
         density_alt = URBRUR) |>
  select(PERSONID,
         HOUSEID,
         year,
         R_AGE,
         ENDTIME,
         TRPMILES,
         mode,
         starts_with("ONTD_P"),
         NUMONTRP,
         NONHHCNT,
         sex,
         HHFAMINC,
         DRVRCNT,
         HHVEHCNT,
         density,
         density_alt) 

#import and filter 2009 trips data
trips09 <- here("nhts",
              "data2009", 
              "DAYV2PUB.csv") |>
  read_csv(show_col_types = FALSE) |>
  mutate(year = 2009) |>
  select (-ONTD_P14,
         -ONTD_P15) |> #removes columns 14 and 15 as these have no data in 2009; allows bind with 2017/22
  filter(WHYFROM == "01" & 
           WHYTO == "21" &
           TRPTRANS %in% c("01",
                           "02",
                           "03",
                           "04",
                           "05",
                           "22",
                           "23")) |>
  mutate(mode = ifelse(TRPTRANS %in% c("22", "23"), "Active", "Car")) |>
  rename(sex = R_SEX,
         density = HBPPOPDN,
         density_alt = URBRUR) |>
  select(PERSONID,
         HOUSEID,
         year,
         R_AGE,
         ENDTIME,
         TRPMILES,
         mode,
         starts_with("ONTD_P"),
         NUMONTRP,
         NONHHCNT,
         sex,
         HHFAMINC,
         DRVRCNT,
         HHVEHCNT,
         density,
         density_alt)

#import and filter 2022 trips data
trips22 <- here("nhts",
                "data2022", 
                "tripv2pub.csv") |>
  read_csv(show_col_types = FALSE) |>
  mutate(year = 2022) |>
  mutate (ONTD_P11 = 0,
         ONTD_P12 = 0,
         ONTD_P13 = 0) |> #add columns to match 2009 and 2017 which go up to person 13)
  filter(WHYFROM %in% c("01","02") & 
           WHYTO == "06" &
           TRPTRANS %in% c("01",
                           "02",
                           "03",
                           "04",
                           "18",
                           "19",
                           "20")) |>
  mutate(mode = ifelse(TRPTRANS %in% c("18", "19","20"), "Active", "Car")) |>
  mutate(density = ifelse(URBRUR == "01", 7000, 50)) |> #estimates continuous density variable for 2022 by assuming rural density = 50 people per sqmile and urban density = 3000 people per sqmile
  rename(sex = R_SEX_IMP,
         density_alt = URBRUR) |> #uses different density variable to 2017 and 2009
  select(PERSONID,
         HOUSEID,
         year,
         R_AGE,
         ENDTIME,
         TRPMILES,
         mode,
         starts_with("ONTD_P"),
         NUMONTRP,
         NONHHCNT,
         sex,
         HHFAMINC,
         DRVRCNT,
         HHVEHCNT,
         density,
         density_alt) 

#combine three files
trips <- rbind(trips09, trips17, trips22) |>
  mutate(arrival = as.numeric(ENDTIME)) |>
  mutate(person_hh_year = paste(PERSONID, HOUSEID, year, sep = "-"),
         hh_year = paste(HOUSEID, year, sep = "-")) |>
  group_by(person_hh_year) |>
  mutate(first_arrival = min(arrival)) |>
  ungroup() |>
  filter(arrival == first_arrival, #only considers first trip of the day (assumed school trip)
         R_AGE > 6 &
         R_AGE < 15 & #only keeps kids trips
         arrival < 1000, #filters trips after 10am, assuming these are school trips
         TRPMILES < 1.25) #filters longer trips
```

Check sample size by year

```{r}
table(trips$year)
```

# Join household member data to trips data

```{r}
trips <- trips |>
  left_join(hh_ages) |>
  left_join(who_in_hh)
```

# Construct variables

We want to include the following variables:

* `log_inc_k`: Income (log-transformed, continuous)
* `veh_per_driver`: Vehicles per driver
* `non_work_mom`: Presence of non-working adult woman in household
* `non_work_dad`: Presence of non-working adult man in household
* `age`: Age
* `female`: Sex
* `has_lil_sib`: Presence of younger child in household
* `has_big_sib`: Presence of older child in household
* `log_distance`: Distance
* `log_density`: Population density of home block group
* `density_alt`: Categorical density variable (urban or rural)

And the outcome variable, `mode_ind`, which can take any of the following values:
    
* 18: Active mode accompanied by a household adult
* 28: Active mode unaccompanied by a household adult
* 17: In a car accompanied by a household adult
* 27: In a car unaccompanied by a household adult

We assume that all of the above alternatives are available for all respondents,
so we will also create the following availability variables, and set the values
to 1/TRUE for all respondents.

* av_kid_car
* av_par_car
* av_kid_act
* av_par_act

## Predictor variables

### Construct income variable

To construct the income variable, we first transform the values to the 
midpoint of the survey category. Then, we inflate/deflate the income to be 
equivalent to the 2017 value (multiply 2009 by 1.14; multiply 2022 by 0.86).

The top income bracket is determined by calculating the highest median income
using census data for the year.

```{r}
trips <- trips |>
  mutate(income_k = case_when(HHFAMINC == "01" & year == 2017 ~ 5,
                              HHFAMINC == "02" & year == 2017 ~ 12.5,
                              HHFAMINC == "03" & year == 2017 ~ 20,
                              HHFAMINC == "04" & year == 2017 ~ 30,
                              HHFAMINC == "05" & year == 2017 ~ 42.5,
                              HHFAMINC == "06" & year == 2017 ~ 62.5,
                              HHFAMINC == "07" & year == 2017 ~ 87.5,
                              HHFAMINC == "08" & year == 2017 ~ 112.5,
                              HHFAMINC == "09" & year == 2017 ~ 137.5,
                              HHFAMINC == "10" & year == 2017 ~ 175,
                              HHFAMINC == "11" & year == 2017 ~ 281.36,
                              HHFAMINC == "01" & year == 2009 ~ 2.5*1.14,
                              HHFAMINC == "02" & year == 2009 ~ 7.5*1.14,
                              HHFAMINC == "03" & year == 2009 ~ 12.5*1.14,
                              HHFAMINC == "04" & year == 2009 ~ 17.5*1.14,
                              HHFAMINC == "05" & year == 2009 ~ 22.5*1.14,
                              HHFAMINC == "06" & year == 2009 ~ 27.5*1.14,
                              HHFAMINC == "07" & year == 2009 ~ 32.5*1.14,
                              HHFAMINC == "08" & year == 2009 ~ 37.5*1.14,
                              HHFAMINC == "09" & year == 2009 ~ 42.5*1.14,
                              HHFAMINC == "10" & year == 2009 ~ 47.5*1.14,
                              HHFAMINC == "11" & year == 2009 ~ 52.5*1.14,
                              HHFAMINC == "12" & year == 2009 ~ 57.5*1.14,
                              HHFAMINC == "13" & year == 2009 ~ 62.5*1.14,
                              HHFAMINC == "14" & year == 2009 ~ 67.5*1.14,
                              HHFAMINC == "15" & year == 2009 ~ 72.5*1.14,
                              HHFAMINC == "16" & year == 2009 ~ 77.5*1.14,
                              HHFAMINC == "17" & year == 2009 ~ 90*1.14,
                              HHFAMINC == "18" & year == 2009 ~ 140*1.14,
                              HHFAMINC == "01" ~ 5*0.86,
                              HHFAMINC == "02" ~ 12.5*0.86,
                              HHFAMINC == "03" ~ 20*0.86,
                              HHFAMINC == "04" ~ 30*0.86,
                              HHFAMINC == "05" ~ 42.5*0.86,
                              HHFAMINC == "06" ~ 62.5*0.86,
                              HHFAMINC == "07" ~ 87.5*0.86,
                              HHFAMINC == "08" ~ 112.5*0.86,
                              HHFAMINC == "09" ~ 137.5*0.86,
                              HHFAMINC == "10" ~ 175*0.86,
                              HHFAMINC == "11" ~ 281.36*0.86,
                              TRUE ~ -9)) |>
  filter(income_k > 0) |>
  mutate(log_inc_k = log(income_k))
```

#### Check: Compare income distribution and means for 2009, 2017, 2022

```{r}

ggplot(trips[trips$year == 2009, ], aes(x = income_k)) +
  geom_histogram(binwidth = 20, fill = "grey", color = "black") +
  ggtitle("Income Distribution in 2009") +
  xlab("Income (in thousands)") +
  ylab("Frequency")

ggplot(trips[trips$year == 2017, ], aes(x = income_k)) +
  geom_histogram(binwidth = 20, fill = "grey", color = "black") +
  ggtitle("Income Distribution in 2017") +
  xlab("Income (in thousands)") +
  ylab("Frequency")

ggplot(trips[trips$year == 2022, ], aes(x = income_k)) +
  geom_histogram(binwidth = 20, fill = "grey", color = "black") +
  ggtitle("Income Distribution in 2022") +
  xlab("Income (in thousands)") +
  ylab("Frequency")

income_2009 <- trips$income_k[trips$year == 2009]
income_2017 <- trips$income_k[trips$year == 2017]
income_2022 <- trips$income_k[trips$year == 2022]

# Compute mean and standard deviation for each sample
summary_stats_income <- aggregate(income_k ~ year, data = trips, FUN = function(x) c(mean = mean(x), sd = sd(x)))

# Print the summary statistics
print.data.frame(summary_stats_income)

```
### Construct vehicles per driver variable

To determine the number of vehicles per driver, we divide the # of household vehicles by the number of drivers in each household. If there are no drivers in the household, we assume a value of 0.

```{r}
trips <- trips |>
  mutate(veh_per_driver = ifelse(DRVRCNT == 0, 0, HHVEHCNT / DRVRCNT))
```

#### Check: Compare vehicle per driver distribution and means for 2009, 2017, 2022

```{r}

ggplot(trips[trips$year == 2009, ], aes(x = veh_per_driver)) +
  geom_histogram(binwidth = 0.5, fill = "grey", color = "black") +
  ggtitle("Vehicles per Driver distribution in 2009") +
  xlab("Vehicles per Driver") +
  ylab("Frequency")

ggplot(trips[trips$year == 2017, ], aes(x = veh_per_driver)) +
  geom_histogram(binwidth = 0.5, fill = "grey", color = "black") +
  ggtitle("Vehicles per Driver distribution in 2017") +
  xlab("Vehicles per Driver") +
  ylab("Frequency")

ggplot(trips[trips$year == 2022, ], aes(x = veh_per_driver)) +
  geom_histogram(binwidth = 0.5, fill = "grey", color = "black") +
  ggtitle("Vehicles per Driver distribution in 2022") +
  xlab("Vehicles per Driver") +
  ylab("Frequency")

vehperdriv_2009 <- trips$veh_per_driver[trips$year == 2009]
vehperdriv_2017 <- trips$veh_per_driver[trips$year == 2017]
vehperdriv_2022 <- trips$veh_per_driver[trips$year == 2022]

# Compute summary statistics
summary_stats_vehperdriv <- aggregate(veh_per_driver ~ year, data = trips, FUN = function(x) c(mean = mean(x), sd = sd(x)))

# Print the summary statistics table
print.data.frame(summary_stats_vehperdriv)

```

### Rename age variable

```{r}
trips <- trips |>
  rename(age = R_AGE)
```

#### Check: Compare age distribution and means for 2009, 2017, 2022

```{r}

ggplot(trips[trips$year == 2009, ], aes(x = age)) +
  geom_histogram(binwidth = 1, fill = "grey", color = "black") +
  ggtitle("Age distribution in 2009") +
  xlab("Age") +
  ylab("Frequency")

ggplot(trips[trips$year == 2017, ], aes(x = age)) +
  geom_histogram(binwidth = 1, fill = "grey", color = "black") +
  ggtitle("Age distribution in 2017") +
  xlab("Age") +
  ylab("Frequency")

ggplot(trips[trips$year == 2022, ], aes(x = age)) +
  geom_histogram(binwidth = 1, fill = "grey", color = "black") +
  ggtitle("Age distribution in 2022") +
  xlab("Age") +
  ylab("Frequency")

age_2009 <- trips$age[trips$year == 2009]
age_2017 <- trips$age[trips$year == 2017]
age_2022 <- trips$age[trips$year == 2022]

# Compute summary statistics
summary_stats_age <- aggregate(age ~ year, data = trips, FUN = function(x) c(mean = mean(x), sd = sd(x)))

# Print the summary statistics table
print.data.frame(summary_stats_age)

```

### Construct female variable

This is derived from the sex variable where 02 = female, 01 = male

```{r}
trips <- trips |>
  mutate(female = ifelse(sex == "02", 1, 0))
```

### Construct non-working parent variables

This is derived from the non-working parent variable constructed from the people file

```{r}
trips <- trips |>
  mutate(non_work_dad = ifelse(non_work_dad, 1, 0),
         non_work_mom = ifelse(non_work_mom, 1, 0))
```

### Construct sibling variables

If an individual's age is not the same as the youngest child in the household, 
they are assumed to have a little sibling (i.e., not be the youngest child); 
if their age is not the same as the oldest child in the household, they are 
assumed to have an older sibling (i.e., they are not the oldest child).

```{r}
trips <- trips |>
  mutate(has_lil_sib = ifelse(age != youngest_kid, 1, 0),
         has_big_sib = ifelse(age != oldest_kid, 1, 0))
```

### Construct distance and density variables

First, we determine the log of continuous distance and density variables (noting the density variable for 2022 are assumed 0, so the density values will be skewed).

```{r}

trips$density <- as.numeric(trips$density) #ensures all values in density variable are numeric

trips <- trips |>
  filter(TRPMILES > 0,
         density > 0) |>
  mutate(log_distance = log(TRPMILES),
         log_density = log(density))
```

#### Check: Compare distance distribution and means for 2009, 2017, 2022

```{r}

ggplot(trips[trips$year == 2009, ], aes(x = TRPMILES)) +
  geom_histogram(binwidth = 0.05, fill = "grey", color = "black") +
  ggtitle("Distance distribution in 2009") +
  xlab("Distance") +
  ylab("Frequency")

ggplot(trips[trips$year == 2017, ], aes(x = TRPMILES)) +
  geom_histogram(binwidth = 0.05, fill = "grey", color = "black") +
  ggtitle("Distance distribution in 2017") +
  xlab("Distance") +
  ylab("Frequency")

ggplot(trips[trips$year == 2022, ], aes(x = TRPMILES)) +
  geom_histogram(binwidth = 0.05, fill = "grey", color = "black") +
  ggtitle("Distance distribution in 2022") +
  xlab("Distance") +
  ylab("Frequency")

distance_2009 <- trips$TRPMILES[trips$year == 2009]
distance_2017 <- trips$TRPMILES[trips$year == 2017]
distance_2022<- trips$TRPMILES[trips$year == 2022]

# Compute summary statistics
summary_stats_distance <- aggregate(TRPMILES ~ year, data = trips, FUN = function(x) c(mean = mean(x), sd = sd(x)))

# Print the summary statistics table
print.data.frame(summary_stats_distance)

```

#### Check: Compare density distribution and means for 2009, 2017, 2022

```{r}

ggplot(trips[trips$year == 2009, ], aes(x = density)) +
  geom_histogram(binwidth = 1000, fill = "grey", color = "black") +
  ggtitle("Density distribution in 2009") +
  xlab("Density") +
  ylab("Frequency")

ggplot(trips[trips$year == 2017, ], aes(x = density)) +
  geom_histogram(binwidth = 1000, fill = "grey", color = "black") +
  ggtitle("Density distribution in 2017") +
  xlab("Density") +
  ylab("Frequency")

ggplot(trips[trips$year == 2022, ], aes(x = density)) +
  geom_histogram(binwidth = 1000, fill = "grey", color = "black") +
  ggtitle("Density distribution in 2022") +
  xlab("Density") +
  ylab("Frequency")

density_2009 <- trips$density[trips$year == 2009]
density_2017 <- trips$density[trips$year == 2017]
density_2022<- trips$density[trips$year == 2022]

# Compute summary statistics
summary_stats_density <- aggregate(density ~ year, data = trips, FUN = function(x) c(mean = mean(x), sd = sd(x)))

# Print the summary statistics table
print.data.frame(summary_stats_density)

```
### Construct year variable

Here, I create two binary variables to determine whether the trip is 2009 or 2017.
All other trips are 2022.

```{r}
trips <- trips |>
  mutate(y2017 = ifelse(year == 2017, 1, 0)) |>
  mutate(y2009 = ifelse(year == 2009, 1, 0))
```

#### Check: Check sample size by year

```{r}
table(trips$year)
```

## Outcome variable

First, we add a variable to indicate whether the trip is accompanied by an adult 
(assumed parent) or not. This is the case if an adult (age>17) is on the trip.

Then, we construct the mode_ind (mode and independence status) for the trip 
according to the following rules:
* 18: Active mode accompanied by a household adult
* 28: Active mode unaccompanied by a household adult
* 17: In a car accompanied by a household adult
* 27: In a car unaccompanied by a household adult

```{r}
trips <- trips |>
  mutate(with_parent = (ONTD_P1 == "01" & age_01 > 17) |
           (ONTD_P2 == "01" & age_02 > 17) |
           (ONTD_P3 == "01" & age_03 > 17) |
           (ONTD_P4 == "01" & age_04 > 17) |
           (ONTD_P5 == "01" & age_05 > 17) |
           (ONTD_P6 == "01" & age_06 > 17) |
           (ONTD_P7 == "01" & age_07 > 17) |
           (ONTD_P8 == "01" & age_08 > 17) |
           (ONTD_P9 == "01" & age_09 > 17) |
           (ONTD_P10 == "01" & age_10 > 17)|
           (ONTD_P11 == "01" & age_11 > 17) |
           (ONTD_P12 == "01" & age_12 > 17) |
           (ONTD_P13 == "01" & age_13 > 17)) |>
  mutate(mode_ind = case_when(with_parent & mode == "Active" ~ 18,
                              with_parent & mode == "Car" ~ 17,
                              mode == "Active" ~ 28, #not with parent
                              mode == "Car" ~ 27, #not with parent
                              TRUE ~ -1))
```

## Availability variables

We assume that all of the above alternatives are available for all respondents,
so we will also create the following availability variables, and set the values
to 1/TRUE for all respondents.

```{r}
trips <- trips |>
  mutate(av_par_car = 1,
         av_par_act = 1,
         av_kid_car = 1,
         av_kid_act = 1)
```

# Select final variable set

```{r}
trips <- trips |>
  select(mode_ind, 
         y2017,
         y2009,
         av_par_car,
         av_par_act,
         av_kid_car,
         av_kid_act,
         log_inc_k,
         veh_per_driver,
         non_work_dad,
         non_work_mom,
         age,
         female,
         has_lil_sib,
         has_big_sib,
         log_distance,
         log_density,
         density_alt)
  
```

# Write csv file

```{r}
write_csv(trips,
          here("models",
               "IATBR plan",
               "trips.csv"))
```

# scale variables 1

```{r}
trips_scaled <- trips |>
  mutate(sc_y2017 = y2017/100,
         sc_log_inc_k = log_inc_k/100,
         sc_age = age/100,
         sc_veh_per_driver = veh_per_driver/100,
         sc_non_work_dad = non_work_dad/100,
         sc_non_work_mom = non_work_mom/100,
         sc_female = female/100,
         sc_has_lil_sib = has_lil_sib/100,
         sc_has_big_sib = has_big_sib/100,
         sc_log_distance = log_distance/100,
         sc_log_density = log_density/100) |>
  select(mode_ind, 
         sc_y2017,
         av_par_car,
         av_par_act,
         av_kid_car,
         av_kid_act,
         sc_log_inc_k,
         sc_veh_per_driver,
         sc_non_work_dad,
         sc_non_work_mom,
         sc_age,
         sc_female,
         sc_has_lil_sib,
         sc_has_big_sib,
         sc_log_distance,
         sc_log_density)
```

# Write csv file

```{r}
write_csv(trips_scaled,
          here("models",
               "IATBR plan",
               "trips_sc.csv"))
```

# scale variables 2

```{r}
trips_scaled2 <- trips |>
  mutate(sc2_y2017 = y2017/10,
         sc2_log_inc_k = log_inc_k/10,
         sc2_age = age/10,
         sc2_veh_per_driver = veh_per_driver/100,
         sc2_non_work_dad = non_work_dad/10,
         sc2_non_work_mom = non_work_mom/10,
         sc2_female = female/10,
         sc2_has_lil_sib = has_lil_sib/10,
         sc2_has_big_sib = has_big_sib/10,
         sc2_log_density = log_density/10) |>
  select(mode_ind, 
         sc2_y2017,
         av_par_car,
         av_par_act,
         av_kid_car,
         av_kid_act,
         sc2_log_inc_k,
         sc2_veh_per_driver,
         sc2_non_work_dad,
         sc2_non_work_mom,
         sc2_age,
         sc2_female,
         sc2_has_lil_sib,
         sc2_has_big_sib,
         log_distance,
         sc2_log_density)
```

# Write csv file

```{r}
write_csv(trips_scaled2,
          here("models",
               "IATBR plan",
               "trips_sc2.csv"))
```

# Create one with combined car alternatives

Here, we combine the two mode_ind involving car travel into a single outcome 
variable (7) because we assume most car trips are accompanied. This results in 
three outcome variables:
* 18: Active mode accompanied by a household adult
* 28: Active mode unaccompanied by a household adult
* 7: In a car (accompanied or unaccompanied)

```{r}
trips3 <- trips |>
  mutate(mode_ind = ifelse(mode_ind %in% c(17, 27), 7, mode_ind ))

write_csv(trips3,
          here("models",
               "IATBR plan",
               "trips3.csv"))
```

# Summaries

```{r}
trips |>
  group_by(y2017) |>
  summarise(car_with_parent = sum(mode_ind == 17),
            car_no_parent = sum(mode_ind == 27),
            active_with_parent = sum(mode_ind == 18),
            active_no_parent = sum(mode_ind == 28))
```


```{r}
trips |>
  group_by(y2017) |>
  summarise(car_with_parent = sum(mode_ind == 17)/n(),
            car_no_parent = sum(mode_ind == 27)/n(),
            active_with_parent = sum(mode_ind == 18)/n(),
            active_no_parent = sum(mode_ind == 28)/n())
```

```{r}
trips |>
  summarise(car_with_parent = sum(mode_ind == 17),
            car_no_parent = sum(mode_ind == 27),
            active_with_parent = sum(mode_ind == 18),
            active_no_parent = sum(mode_ind == 28))
```


```{r}
trips |>
  summarise(car_with_parent = sum(mode_ind == 17)/n(),
            car_no_parent = sum(mode_ind == 27)/n(),
            active_with_parent = sum(mode_ind == 18)/n(),
            active_no_parent = sum(mode_ind == 28)/n())
```

