---
title: "IATBR data assembly"
output: html_document
date: "2024-04-01"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Required libraries

We're only using the `tidyverse` and `here` libraries.

```{r, message=FALSE}
library(tidyverse)
library(here)
```

# Assemble person-level data

We need information on the ages of the household members on each trip to 
determine whether a child was accompanied by a household adult.

We also need to determine whether each household includes a non-working female
adults and/or a non-working male adult and the ages of the youngest and 
oldest children in each household (to determine if the trip-maker has a 
younger or older sibling).

## Load person-level data

We start by loading the person-level data files for both years, selecting the
same set of variables from each year, adding a variable to indicate which
year's sample the household is from, and renaming the sex variable to have the
same name in both years.

Then we combine the two datasets. 

```{r}
people17 <- here("nhts",
                 "data2017", 
                 "perpub.csv") |>
  read_csv(show_col_types = FALSE) |>
  select(HOUSEID, 
         PERSONID,
         R_AGE,
         R_SEX_IMP,
         WORKER,
         HHSIZE) |>
  mutate(year = 2017) |>
  rename(sex = R_SEX_IMP)

people09 <- here("nhts",
               "data2009", 
               "PERV2PUB.csv") |>
  read_csv(show_col_types = FALSE) |>
  select(HOUSEID, 
         PERSONID,
         R_AGE,
         R_SEX,
         WORKER,
         HHSIZE) |>
  mutate(year = 2009) |>
  rename(sex = R_SEX)

people <- rbind(people09, people17) |>
  mutate(person_hh_year = paste(PERSONID, HOUSEID, year, sep = "-")) |>
  mutate(hh_year = paste(HOUSEID, year, sep = "-"))

```

## Get ages of household members

We'll reshape the person-level data into a household-level dataset with
a column indicating the age of each person in the household.

```{r}
hh_ages <- people |>
  select(hh_year, PERSONID, R_AGE) |>
  pivot_wider(names_from = PERSONID,
              names_prefix = "age_",
              values_from = R_AGE)
```

## Get presence of non-working adult men and women and age ranges of children

```{r}
who_in_hh <- people |>
  mutate(adult = R_AGE > 17,
         kid_age = ifelse(R_AGE < 18, R_AGE, -2),
         R_AGE = ifelse(R_AGE < 0, 999, R_AGE),
         non_worker_mom = R_AGE > 17 & WORKER != "01" & sex == "02",
         non_worker_dad = R_AGE > 17 & WORKER != "01" & sex != "02") |>
  group_by(hh_year, HHSIZE) |>
  summarize(num_records = n(),
         n_adults = sum(adult),
         non_work_mom = sum(non_worker_mom) > 0,
         non_work_dad = sum(non_worker_dad) > 0,
         min_age = min(R_AGE),
         max_kid_age = max(kid_age)) |>
  mutate(youngest_kid = ifelse(num_records < HHSIZE, 1, min_age),
         oldest_kid = ifelse(max_kid_age > 0, max_kid_age, 1),
         n_children = HHSIZE - n_adults) |>
  ungroup() |>
  filter(n_children > 0) |>
  select(hh_year, 
         non_work_dad,
         non_work_mom,
         oldest_kid, 
         youngest_kid)
```

## Trips data

```{r}
trips17 <- here("nhts",
                "data2017", 
                "trippub.csv") |>
  read_csv(show_col_types = FALSE) |>
  mutate(year = 2017) |>
  filter(WHYFROM %in% c("01","02") & 
           WHYTO == "08" &
           TRPTRANS %in% c("01",
                           "02",
                           "03",
                           "04",
                           "05",
                           "06")) |>
  mutate(mode = ifelse(TRPTRANS %in% c("01", "02"), "Active", "Car")) |>
  rename(sex = R_SEX_IMP,
         density = OBPPOPDN) |>
  select(PERSONID,
         HOUSEID,
         year,
         R_AGE,
         ENDTIME,
         TRPMILES,
         mode,
         starts_with("ONTD_P"),
         NUMONTRP,
         NONHHCNT,
         sex,
         HHFAMINC,
         DRVRCNT,
         HHVEHCNT,
         density) 

trips09 <- here("nhts",
              "data2009", 
              "DAYV2PUB.csv") |>
  read_csv(show_col_types = FALSE) |>
  mutate(year = 2009) |>
  select(-ONTD_P15, -ONTD_P14) |>
  filter(WHYFROM == "01" & 
           WHYTO == "21" &
           TRPTRANS %in% c("01",
                           "02",
                           "03",
                           "04",
                           "05",
                           "22",
                           "23")) |>
  mutate(mode = ifelse(TRPTRANS %in% c("22", "23"), "Active", "Car")) |>
  rename(sex = R_SEX,
         density = HBPPOPDN) |>
  select(PERSONID,
         HOUSEID,
         year,
         R_AGE,
         ENDTIME,
         TRPMILES,
         mode,
         starts_with("ONTD_P"),
         NUMONTRP,
         NONHHCNT,
         sex,
         HHFAMINC,
         DRVRCNT,
         HHVEHCNT,
         density)

trips <- rbind(trips09, trips17) |>
  mutate(arrival = as.numeric(ENDTIME)) |>
  mutate(person_hh_year = paste(PERSONID, HOUSEID, year, sep = "-"),
         hh_year = paste(HOUSEID, year, sep = "-")) |>
  group_by(person_hh_year) |>
  mutate(first_arrival = min(arrival)) |>
  ungroup() |>
  filter(arrival == first_arrival,
         R_AGE > 6 &
         R_AGE < 15 & 
         arrival < 1000,
         TRPMILES < 1.25) 
```

Check sample size by year

```{r}
table(trips$year)
```

# Add household member data to trips data

```{r}
trips <- trips |>
  left_join(hh_ages) |>
  left_join(who_in_hh)
```

# Construct variables

We want to include the following variables:

* `log_inc_k`: Income (log-transformed, continuous)
* `veh_per_driver` :Vehicles per driver
* `non_work_mom`: Presence of non-working adult woman in household
* `non_work_dad`: Presence of non-working adult man in household
* `age`: Age
* `female`: Sex
* `has_lil_sib`: Presence of younger child in household
* `has_big_sib`: Presence of older child in household
* `log_distance`: Distance
* `log_density`: Population density of home block group

And the outcome variable, `mode_ind`, which can take any of the following values:
    
* 18: Active mode accompanied by a household adult
* 28: Active mode unaccompanied by a household adult
* 17: In a car accompanied by a household adult
* 27: In a car unaccompanied by a household adult

We assume that all of the above alternatives are available for all respondents,
so we will also create the following availability variables, and set the values
to 1/TRUE for all respondents.

* av_kid_car
* av_par_car
* av_kid_act
* av_par_act

## Predictor variables

### Construct income variable

```{r}
trips <- trips |>
  mutate(income_k = case_when(HHFAMINC == "01" & year == 2017 ~ 5,
                              HHFAMINC == "02" & year == 2017 ~ 12.5,
                              HHFAMINC == "03" & year == 2017 ~ 20,
                              HHFAMINC == "04" & year == 2017 ~ 30,
                              HHFAMINC == "05" & year == 2017 ~ 42.5,
                              HHFAMINC == "06" & year == 2017 ~ 62.5,
                              HHFAMINC == "07" & year == 2017 ~ 87.5,
                              HHFAMINC == "08" & year == 2017 ~ 112.5,
                              HHFAMINC == "09" & year == 2017 ~ 137.5,
                              HHFAMINC == "10" & year == 2017 ~ 175,
                              HHFAMINC == "11" & year == 2017 ~ 281.36,
                              HHFAMINC == "01" ~ 2.5*1.14,
                              HHFAMINC == "02" ~ 7.5*1.14,
                              HHFAMINC == "03" ~ 12.5*1.14,
                              HHFAMINC == "04" ~ 17.5*1.14,
                              HHFAMINC == "05" ~ 22.5*1.14,
                              HHFAMINC == "06" ~ 27.5*1.14,
                              HHFAMINC == "07" ~ 32.5*1.14,
                              HHFAMINC == "08" ~ 37.5*1.14,
                              HHFAMINC == "09" ~ 42.5*1.14,
                              HHFAMINC == "10" ~ 47.5*1.14,
                              HHFAMINC == "11" ~ 52.5*1.14,
                              HHFAMINC == "12" ~ 57.5*1.14,
                              HHFAMINC == "13" ~ 62.5*1.14,
                              HHFAMINC == "14" ~ 67.5*1.14,
                              HHFAMINC == "15" ~ 72.5*1.14,
                              HHFAMINC == "16" ~ 77.5*1.14,
                              HHFAMINC == "17" ~ 90*1.14,
                              HHFAMINC == "18" ~ 140*1.14,
                              TRUE ~ -9)) |>
  filter(income_k > 0) |>
  mutate(log_inc_k = log(income_k))
```

### Construct vehicles per driver variable

```{r}
trips <- trips |>
  mutate(veh_per_driver = ifelse(DRVRCNT == 0, 0, HHVEHCNT / DRVRCNT))
```

### Rename age variable

```{r}
trips <- trips |>
  rename(age = R_AGE)
```

### Construct female variable

```{r}
trips <- trips |>
  mutate(female = ifelse(sex == "02", 1, 0))
```

### Construct non-working parent variables

```{r}
trips <- trips |>
  mutate(non_work_dad = ifelse(non_work_dad, 1, 0),
         non_work_mom = ifelse(non_work_mom, 1, 0))
```

### Construct sibling variables

```{r}
trips <- trips |>
  mutate(has_lil_sib = ifelse(age != youngest_kid, 1, 0),
         has_big_sib = ifelse(age != oldest_kid, 1, 0))
```

### Construct distance and density variables

```{r}
trips <- trips |>
  filter(TRPMILES > 0,
         density > 0) |>
  mutate(log_distance = log(TRPMILES),
         log_density = log(density))
```

### Construct year variable

```{r}
trips <- trips |>
  mutate(y2017 = ifelse(year == 2017, 1, 0))
```


### Check sample size by year

```{r}
table(trips$year)
```
## Outcome variable

```{r}
trips <- trips |>
  mutate(with_parent = ONTD_P1 == "01" & age_01 > 17 |
           ONTD_P2 == "01" & age_02 > 17 |
           ONTD_P3 == "01" & age_03 > 17 |
           ONTD_P4 == "01" & age_04 > 17 |
           ONTD_P5 == "01" & age_05 > 17 |
           ONTD_P6 == "01" & age_06 > 17 |
           ONTD_P7 == "01" & age_07 > 17 |
           ONTD_P8 == "01" & age_08 > 17 |
           ONTD_P9 == "01" & age_09 > 17 |
           ONTD_P10 == "01" & age_10 > 17 |
           ONTD_P11 == "01" & age_11 > 17 |
           ONTD_P12 == "01" & age_12 > 17 |
           ONTD_P13 == "01" & age_13 > 17) |>
  mutate(mode_ind = case_when(with_parent & mode == "Active" ~ 18,
                              with_parent & mode == "Car" ~ 17,
                              mode == "Active" ~ 28,
                              mode == "Car" ~ 27,
                              TRUE ~ -1))
```

## Availability variables

```{r}
trips <- trips |>
  mutate(av_par_car = 1,
         av_par_act = 1,
         av_kid_car = 1,
         av_kid_act = 1)
```

# Select final variable set

```{r}
trips <- trips |>
  select(mode_ind, 
         y2017,
         av_par_car,
         av_par_act,
         av_kid_car,
         av_kid_act,
         log_inc_k,
         veh_per_driver,
         non_work_dad,
         non_work_mom,
         age,
         female,
         has_lil_sib,
         has_big_sib,
         log_distance,
         log_density)
  
```

# Write csv file

```{r}
write_csv(trips,
          here("models",
               "IATBR plan",
               "trips.csv"))
```

