---
title: "Survey Quintile Example"
output: html_document
date: "2024-03-14"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Required packages

This example uses the following packages:

```{r, message=FALSE}
library(tidyverse)
library(srvyr)
library(here)
library(knitr)
```

# Load data and create survey object

Here, I'm just using the household data from the 2017 NHTS.

In one long pipe, I'm: 

* Reading in the data
* Creating a continuous version fo the income variable
* Filtering rows with missing income values
* Converting this data frame to a survey object 
* Calculating quintile thresholds
* Comparing the continuous income value to the threshold value to deterime 
which income quintile each household is in.

```{r, message=FALSE}
hh_2017 <- here("nhts",
                "data2017",
                "hhpub.csv") |>
  # read data
  read_csv() |> 
  # make numeric income variable based on category midpoints
  mutate(income_k = case_when(HHFAMINC == "01" ~ 5, 
                              HHFAMINC == "02" ~ 12.5,
                              HHFAMINC == "03" ~ 20,
                              HHFAMINC == "04" ~ 30,
                              HHFAMINC == "05" ~ 42.5,
                              HHFAMINC == "06" ~ 62.5,
                              HHFAMINC == "07" ~ 87.5,
                              HHFAMINC == "08" ~ 112.5,
                              HHFAMINC == "09" ~ 137.5,
                              HHFAMINC == "10" ~ 175,
                              HHFAMINC == "11" ~ 250,
                              TRUE ~ -9)) |>
  # filter rows with missing value for income
  filter(income_k > 0) |> 
  # create survey object with weights
  as_survey_design(weights = WTHHFIN) |> 
  # Calculate quintile thresholds
  mutate(quint_break = survey_quantile(income_k, 
                                       quantiles = c(0.2, 
                                                     0.4, 
                                                     0.6, 
                                                     0.8))) |>
  # compare income_k value to thresholds to get income quintile of each HH
  mutate(inc_quint = case_when(income_k < quint_break$`_q20`[1] ~ "q1",
                               income_k < quint_break$`_q40`[1] ~ "q2",
                               income_k < quint_break$`_q60`[1] ~ "q3",
                               income_k < quint_break$`_q80`[1] ~ "q4",
                               TRUE ~ "q5"))
```

# Check quintile proportions

To see how this all worked, we can create a summary table showing the 
(unweighted) number of respondents in each income category.

```{r}
hh_2017 |>
  group_by(inc_quint) |>
  summarise(n = n()) |>
  mutate(proportion = n/sum(n)) |>
  kable(digits = c(0, 0, 2))
```

Based on the unweighted numbers, we see that only 10 percent of households
are in the bottom quintile. There are two reasons for this: 

1. These are unweighted counts
2. All households in the same NHTS income category are given the same 
assumed income. If this value is right on the threshold, they will all
fall on the same side.

Now let's see the weighted counts.

```{r}
hh_2017 |>
  group_by(inc_quint) |>
  survey_tally(vartype = "ci") |>
  mutate(proportion = n/sum(n)) |>
  kable(digits = c(0, 0, 0, 0, 2))
```

Even with the weights applied, we still don't have even proportions in each 
category - although it's closer now with 14 rather than 10 percent in the lowest
quintile. 

# Recalculate quintiles with new threshold rule

We assigned households into quintiles by saying those with incomes below the
threshold would fall into the lower quintile. Let's try it again with a rule that
households with incomes less than _or equal to_ the threshold go into the 
lower quintile.

```{r}
hh_2017 <- hh_2017 |>
  # compare income_k value to thresholds to get income quintile of each HH
  mutate(inc_quint_2 = case_when(income_k <= quint_break$`_q20`[1] ~ "q1",
                                 income_k <= quint_break$`_q40`[1] ~ "q2",
                                 income_k <= quint_break$`_q60`[1] ~ "q3",
                                 income_k <= quint_break$`_q80`[1] ~ "q4",
                                 TRUE ~ "q5"))
```


# Recheck proportions

Now let's see how the unweighted proportions look.

```{r}
hh_2017 |>
  group_by(inc_quint_2) |>
  summarise(n = n()) |>
  mutate(proportion = n/sum(n)) |>
  kable(digits = c(0, 0, 2))
```

And here are the weighted proportions:

```{r}
hh_2017 |>
  group_by(inc_quint_2) |>
  survey_tally(vartype = "ci") |>
  mutate(proportion = n/sum(n)) |>
  kable(digits = c(0, 0, 0, 0, 2))
```


These proportions are much closer to having 20 percent of all households in 
each quintile.

It's interesting that the unweighted proportions are much closer to even
quintiles than the weighted proportions -- so this doesn't work quite as well
as I might have hoped as a demonstration of the importance of weighting. They're
both pretty close though, and weighting is still important when you calculate
summary statistics that are meant to represent the population you've sampled
from.